### 2013-02-04 ElsevierFGCS_singleColumn.tex
FGCS

\documentclass[sort&compress,preprint,3p]{elsarticle}
\usepackage{color}
\usepackage{url}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{listings}
\usepackage{enumerate}
\usepackage{textcomp}
\usepackage[caption=false,font=footnotesize]{subfig}
\usepackage[table]{xcolor}
\definecolor{light-gray}{gray}{0.80}
\DeclareMathSizes{8}{8}{8}{8}

\usepackage{amssymb}
\journal{Future Generation Computer Systems}

\begin{document}
\begin{frontmatter}
\title{Using Model Checking to Analyze the System Behavior of the LHC Production Grid}

\author[VU,NIKHEF]{Daniela Remenska\corref{cor1}}
\ead{danielar@nikhef.nl}
\author[TUe]{Tim A.C.Willemse}
\ead{timw@win.tue.nl}
\author[VU]{Kees Verstoep}
\ead{c.verstoep@vu.nl}
\author[NIKHEF]{Jeff Templon}
\ead{templon@nikhef.nl}
\author[VU]{Henri Bal}
\ead{bal@cs.vu.nl}
\address[VU]{Department of Computer Science, VU University Amsterdam, The Netherlands}
\address[TUe]{Department of Computer Science, TU Eindhoven, The Netherlands}
\address[NIKHEF]{NIKHEF National Institute for High-Energy Physics, Amsterdam, The Netherlands}

\cortext[cor1]{Corresponding author}

\begin{abstract}
DIRAC (Distributed Infrastructure with Remote Agent Control) is the grid
solution designed to support production activities as well as user data analysis
for the Large Hadron Collider ``beauty'' experiment. It consists of cooperating
distributed services and a plethora of light-weight agents delivering the
workload to the grid resources.  Services accept requests from agents and
running jobs, while agents actively fulfill specific goals. Services maintain
database back-ends to store dynamic state information of entities such as jobs,
queues, or requests for data transfer. Agents continuously check for changes in the
service states, and react to these accordingly. The logic of each agent is
rather simple; the main source of complexity lies in their cooperation. These
agents run concurrently, and communicate using the services' databases as a
shared memory for synchronizing the state transitions. Despite the effort
invested in making DIRAC reliable, entities occasionally get into inconsistent
states. Tracing and fixing such behaviors is difficult, given the inherent
parallelism among the distributed components and the size of the
implementation.

In this paper we present an analysis of DIRAC with mCRL2, process algebra with
data. We have reverse engineered two critical and related DIRAC subsystems, and
subsequently modeled their behavior with the mCRL2 toolset. This enabled us to
easily locate race conditions and livelocks which were confirmed to occur in the
real system. We further formalized and verified several behavioral properties of
the two modeled subsystems.
\end{abstract}

\begin{keyword}

model checking \sep process algebra \sep grid \sep LHC \sep distributed system \sep workflow
\end{keyword}

\end{frontmatter}

\section{Introduction}

The Large Hadron Collider beauty (LHCb) experiment \cite{LHCb} is one of the
four large
experiments conducted on the Large Hadron Collider (LHC) accelerator, built by
the European Organization for Nuclear Research (CERN). Immense amounts of data
are produced at the LHC accelerator, and subsequently processed by physics
groups and individuals worldwide. The sheer size of the experiment is the
motivation behind the adoption of the grid computing paradigm. The grid storage
and computing resources for the LHCb experiment are distributed across several
institutes in Europe. To cope with the complexity of processing the vast amount
of data, a complete grid solution, called DIRAC (Distributed Infrastructure with
Remote Agent Control) \cite{DIRAC_CommGridSolution,DIRAC_ReliableDataMaangement}, 
has been designed and developed for the
LHCb
community.

DIRAC forms a layer between the LHCb user community and the heterogeneous
grid resources, to allow for optimal and reliable usage of these
resources. It consists of many cooperating distributed services and light-weight
agents which deliver the workload to the resources. 
Services maintain database back-ends to store
dynamic state information of entities such as jobs, queues, staging requests, etc. Agents use polling to check
and possibly react to changes in the service states.
The logic of each individual
component is relatively simple; the overall system complexity emerges from the
cooperation among them. Namely, these agents run concurrently, and communicate
using the services' databases as a shared memory (blackboard paradigm \cite{IEEEexample:blackboard_systems}) for
synchronizing state transitions of various entities.

Although much effort is invested in making DIRAC reliable, entities occasionally
get into inconsistent states, leading to a loss of efficiency in both
resource usage and manpower. Debugging and fixing the root of such encountered
behaviors becomes a formidable mission due to multiple factors: the inherent
parallelism present among the system components deployed on different physical
machines, the size of the implementation ($\sim$150,000 lines of Python code),
and the distributed knowledge of different subsystems within the collaboration.

In this paper we propose the use of rigorous (formal) methods for improving
software quality. Model checking \cite{ProcessesWithData} is one such technique for analysis of an
abstract model of a system, and verification of certain system properties of
interest. Unlike conventional testing, it allows full control over the execution
of parallel processes and also supports automated exhaustive state-space
exploration. We used the mCRL2 language \cite{FormalLanguagemCRL2} and toolset \cite{mCRL2Toolset} to model the behavior of two
critical and related DIRAC components: the Workload Management (WMS) and the
Storage Management System (SMS). Based on Algebra of Communicating Processes
(ACP) \cite{process_algebra}, mCRL2 is able to deal with generic data types as well as user-defined
functions for data transformation. This makes it particularly suitable for
modeling the data manipulations made by DIRAC's agents. Visualizing the state
space and replaying scenarios with the toolkit's simulator enabled us to gain
insight into the system behavior and incrementally improve the model. 
Already using these techniques, critical race conditions and livelocks were detected and confirmed to
occur in the real system. Some of them were a result of simple coding bugs;
others unveiled more elementary design problems. We further formulated,
formalized and verified several application-specific properties.

The idea of modeling existing systems using 
formal techniques is as such not new. Earlier studies
(\cite{SPIN_case_study,sessionMgmtMasterThesis,desing_validation_protocols,protocol_verification_muCRL,SLAMToolkit,SlidingWindowProtocol,DHCP_SPIN}) mostly focused
on modeling and verifying hardware or communication protocols, since
the formal languages and tools at hand were not sufficiently mature
to cope with more complex data-intensive distributed systems. More
recently, success stories on modeling real-life concurrent systems with
data have been reported (\cite{CMS_LHC, Linux_driver,verstoep_et_al, SystemC_processAlgebra,Java_PathFinder, ACS_mCRL2}).
Tools for automatic translation of the system implementation language into
 formal language constructs can greatly
simplify the analysis. However, this has so far been feasible only when the
language of implementation is domain-specific, or alternatively, a
reasonably small subset of a general-purpose language is considered for
translation.

We believe that the challenges and results of this work are unique in
a number of aspects. First, to the best of our knowledge, the code-base
and the number of concurrent grid components engaged in providing DIRAC's
functionality considerably outnumber most of the previous industrial cases. Second,
the choice of Python as implementation platform has led to the prevailing
usage of dynamic structures (whose types and sizes are determined at
runtime) throughout DIRAC, challenging the transition to an abstract
formal representation. We have established general
guidelines on extracting a model outline from the implementation. These
can be reused in many distributed systems to address concurrency issues arising 
from the use of shared storage for inter-process communication. Third,
analysis of this kind is typically performed after a problem has already
surfaced in the real system, as a means to understand the events which
led to it and test for possible solutions. We managed to come across an
actual bug at the same time it was observed in practice, which increased
our confidence in the soundness of the model.

The paper is organized as follows. Section 2 introduces the architecture
of DIRAC, focusing on the two subsystems chosen as case studies. Section 3 gives a brief
overview of the mCRL2 language, and describes our approach to abstracting
and modeling the behavior of these subsystems. Section 4 presents the
analysis with the mCRL2 toolset and the issues detected. Section 5 and 6
conclude and give directions for future work, respectively.

\section{DIRAC: A Community Grid Solution}
\label{sec:Section_2}
\subsection{Architecture Overview}

The development of DIRAC started in 2002 as a system for production of simulation
data that would serve to verify physics theory, aspects of the LHCb detector design, as
well as to optimize physics algorithms. It gradually evolved into a complete community
grid solution for data and job management, based on a general-purpose framework
that can be reused by other communities besides LHCb. Today, it covers all major
LHCb tasks starting with the raw data transfer from the experiment's detector to
the grid storage, several steps of data processing, up to the final user
analysis. 
It provides a concurrent use of over 60K CPUs and 10M file replicas distributed over
tens of grid sites, provided by the WLCG computing grid \cite{Bird:2011zz}. The community of users has grown to several hundreds, loading the 
LHC grid resources with up to 30K simultaneously running jobs (Fig.~\ref{fig:LHCb-resource-usage}, top) and over 100000 jobs executed per day, during peak processing periods.
About 10PB of data is stored and visible in the LHCb file catalogs, with daily data transfers among storage units reaching 3Gbps (Fig.~\ref{fig:LHCb-resource-usage}, bottom).
\begin{figure*}[tp!]
  \centering
  {\label{fig:Job-throughput}\includegraphics[width=0.7\linewidth]{./graphics/Figure1_1.png}}    
  \vfill
  \vspace{14pt}
{\label{fig:Data-transfers}\includegraphics[width =0.7\linewidth]{./graphics/Figure1_2.png}}
  \caption{LHCb resource usage}
  \label{fig:LHCb-resource-usage}
\end{figure*}
Python was chosen as the implementation language, since it enables
rapid prototyping and development of new features. DIRAC follows the Service
Oriented Architecture (SOA) paradigm, accompanied by a network of lightweight
distributed agents which animate the system. Its main components are depicted in
Fig.~\ref{fig:DIRAC-Arch}.

\begin{figure}[tp]
\includegraphics[width=0.5\textwidth] {./graphics/Figure1.png}
\centering
\caption{DIRAC Architecture overview}
\label{fig:DIRAC-Arch}
\end{figure}

The \textit{services} are passive components that react to requests from their clients,
possibly soliciting other services in order to fulfill the requests. They run as
permanent processes deployed on a number of high-availability hosts (VO-boxes)
at CERN, and store the dynamic system state information in database
repositories. The user interfaces, agents or running jobs can act as clients
placing the requests to DIRAC's services.

\textit{Agents} are active components that fulfill a limited number of specific system
functions. They run in different environments, depending on their mission.
Some are deployed close to the corresponding services, while others run on the
grid worker nodes. Examples of the latter are the so-called Pilot Agents, part
of the Workload Management System explained in the following section. All DIRAC
agents repeat the same logic in each iteration cycle: they monitor the service states,
 and respond by initiating actions (like job
submission or data transfer) which update the states of various system
entities.

\textit{Resources} are software abstractions of the underlying heterogeneous grid
computing and storage entities allocated to LHCb, providing a uniform interface
for access. The physical resources are controlled by the site managers and made
available through middleware services such as gLite \cite{Laure06programmingthe}.

The DIRAC functionality is exposed to users and developers through a rich set of
command-line tools forming the DIRAC API, complemented by a Web portal for
visually monitoring  the system behavior and controlling the ongoing tasks. Both
the Web and command-line \textit{interfaces} ensure secure system access using X509
certificates.

The DIRAC hardware configuration of LHCb is still by far the most powerful one: it consists of about 10 servers (VO-boxes) located at CERN, hosting the central
DIRAC services, including 6 MySQL database servers. Small fraction of the servers and databases are used for testing purposes, and are connected
to the same pool of WLCG computing and storage resources. Each of the six LHCb Tier-1 computing centers accross Europe hosts an additional server as part of this cluster for redundancy,
to ensure high service availability at any moment. Several other experiments are currently using DIRAC on an every day basis, 
with new user groups evaluating its functionality and performance before adopting it in production.

In the following, we focus on two related subsystems that are considered the backbone of DIRAC. These
are the ones where problematic state changes are most often encountered,
 which are difficult to trace and correct. Understanding their behavior 
is essential for interpreting the issues that were discovered during the model analysis and verification.

\subsection{Workload Management System}

The driving force of DIRAC is the Workload Management System (WMS). Taking into
account the heterogeneous, dynamic, and high-latency nature of the distributed computing environment, a
\textit{Pilot Job paradigm}, illustrated in Fig.~\ref{fig:DIRAC-WMS}, 
\begin{figure}[bp]
\includegraphics[width=0.7\textwidth]{./graphics/Figure2.png}
\centering
\caption{DIRAC Workload Management System \cite{DIRAC_pilot_WMS}}
\label{fig:DIRAC-WMS}
\end{figure}
was chosen as an efficient way to
implement a pull scheduling mechanism.  Pilot jobs are simply resource
reservation processes without an actual payload defined a priori. They are
submitted to the grid worker nodes with the aim of checking the sanity of the
operational environment just before pulling and executing the real payload. This
hides the fragility of the underlying resources and increases the job success
rate, from the perspective of end users. Centralized \textit{Task Queues} contain all the
pending jobs, organized in groups according to their requirements. This enables
efficient application of job priorities, enforcing fair-share policies and
quotas, which are decided within the LHCb community, without the need of any
effort from the grid sites.  Originally, the DIRAC WMS submitted pilot jobs using 
the native gLite middleware tools, but this method was shown to be suboptimal. The gLite resource brokers
are centralized components, and as such they do not have the capacity and flexibility to cope with the load of 
a community such as LHCb. Using multiple resource brokers, rather than a single one, was not the solution:
all brokers use the same site state information and would always submit jobs to the most suitable one globally, causing overloads to
certain sites while underloading others very often. In addition, the gLite information system was not reactive enough to propagate
changes in site status information to the brokers. With the wide deployment of the CREAM CE service \cite{Aiftimiei2010654}, DIRAC's WMS switched
to submitting pilot jobs directly to sites. By interrogating the Computing Element status directly, the system reactiveness
improved dramatically, making sites compete with each other for jobs and have a faster job turnaround.

The basic flowchart describing the evolution of a job's states is depicted in
Fig.~\ref{fig:DIRAC-job-state-machine}. After submission through the \textit{Job Manager} (``Received''), the complete
job description is placed in the DIRAC job repository (the Job DB). Before jobs
become eligible for execution, a chain of optimizer agents checks and prioritizes
them in queues, utilizing the parameters information from the Job DB.
\begin{figure}[tp]
\includegraphics[width=0.7\linewidth,keepaspectratio=true]{./graphics/Figure3.png}
\centering
\caption{Job state machine \cite{ProductionShifterGuide}}
\label{fig:DIRAC-job-state-machine}
\end{figure} 
If the requested data resides on tape storage, the \textit{Job
Scheduling Agent} will pass the control to a specialized \textit{Stager} service (part of
the SMS explained in the next section), before placing the job in a Task Queue
(``Waiting''). Based on the complete list of pending payloads, a specialized \textit{Task
Queue Director} agent submits pilots to the computing resources via the available job
submission middleware (e.g., gLite WMS). After
a \textit{Matcher} service pulls the most suitable payload for a pilot (``Matched''), a \textit{Job
Wrapper} object is created on the worker node, responsible for retrieving the
input sandbox, performing software availability checks, executing the actual
payload on the worker node (``Running''), and finally uploading any output data
necessary (``Done'' or ``Failed''). The wrapper can catch any failure exit state
reported by the running physics applications. At the same time, a \textit{Watchdog}
process is instantiated to monitor the behavior of the Job Wrapper and send
heartbeat signals to the monitoring service. It can also take actions in case
resources are soon to be exhausted, the payload stalls, or a management command
for killing the payload is received.

Although the grid storage resources are limited, it is essential to keep all
data collected throughout the experiment's run. Tape backends provide a reliable
and cheap solution for data storage. The additional workflow step 
necessary for input data files residing on tape is carried out inside the
Storage Management System (SMS).

\subsection{Storage Management System}

The DIRAC SMS provides the logic for pre-staging files from tape to a disk cache
frontend, before a job is able to process them. Smooth functioning of this
system is essential for production activities which involve reprocessing 
older data with improved physics software, and happens typically several times
per year.

A simplified view of the system is shown in  Fig.~\ref{fig:DIRAC-SMS}. The workflow is initiated
with the \textit{Job Scheduling Agent} detecting that a job is assigned to process files
only available on tape storage. It sends a request for staging (i.e., creating a
cached replica) to the \textit{Storage Manager Handler} service with the
list of files and a callback method to be invoked when the request has been
processed. The Storage Management DB is immediately populated with records which
are processed by a sequence of agents in an organized fashion. The relevant
tables in the SMS DB are the \textit{Tasks} and \textit{CacheReplicas}, whose
entities maintain a state observed and updated by these agents. \textit{Tasks} maintain
general information about every job requesting a service from the SMS. The
details about every file (i.e., the Storage Element where it resides, the
size, checksum, number of tasks that requested it), are kept in the \textit{CacheReplicas}
table. Other auxiliary tables maintain the relationship between these entities.
\begin{figure}[tp]
\begin{center}
 \includegraphics[width=0.5\linewidth,keepaspectratio=true]{./graphics/Figure4.png}
\end{center}
\caption{DIRAC Storage Management System}
\label{fig:DIRAC-SMS}
\end{figure}
The processing begins with the \textit{Request Preparation Agent}. It selects all the
``New'' replica entries, checks whether they are registered in a Logical
File Catalog (LFC), and retrieves their metadata. In case of problematic catalog
entries, it can update the state of the \textit{CacheReplicas} and the related \textit{Tasks}
entries to ``Failed''. Non-problematic files are updated to a
``Waiting'' state. The \textit{Stage Request Agent} is responsible for placing the actual staging
requests for all ``Waiting'' entries, via dedicated storage middleware that
communicates with the tape backends. These requests are grouped by Storage Element (SE) prior to
submission, and carry information about the requested (pin) lifetime of the
replicas to be cached. If certain pathologies are discovered (i.e., lost,
unavailable, or zero-sized files on tape), it can update the corresponding
entries to ``Failed'' in a similar manner. Otherwise, they are promoted to
``StageSubmitted''. The agent responsible for monitoring the status of submitted
requests is the \textit{Stage Monitor Agent}. It achieves this by interrogating the
storage middleware to see if the ``StageSubmitted'' files are successfully
replicated on disk cache. In case of success, the \textit{CacheReplicas} and their
corresponding \textit{Tasks} entries are updated to ``Staged''. Various circumstances of
tape or middleware misbehavior can also fail the staging requests. The last one
in the chain is the \textit{Request Finalization Agent}. The \textit{Tasks} which are in their
final states (``Staged'' or ``Failed'') are cleared from the database, and callbacks
are performed to the WMS, which effectively wakes up the corresponding jobs. If
there are no more associated \textit{Tasks} for particular replicas, the respective
\textit{CacheReplicas} entries are also removed.

This subsystem is still in an evolutionary phase. In multiple instances, tasks or
replicas have become “stuck”, effectively blocking the progress of jobs. Tracing
back the sequence of events which led to the inconsistent states is non-trivial.
To temporarily alleviate such problems, the status of these entries is typically manually
reset to the initial ``New'' state, so that agents can re-process them from
scratch. Occasionally, error messages are reported from unsuccessful attempts of
the SMS service to update the state of non-existent table entries.

\section{System modeling}

\subsection{The mCRL2 language and toolset}

Distributed systems, such as the DIRAC system, are commonly modeled
by a directed, edge-labeled graph referred to as a \emph{Labeled
Transition System} (LTS). The nodes in the graph represent the states
of the system, edges between nodes represent an atomic state change
and the edge labelings represent atomic events such as reading, sending
and successful communications within the system. \emph{Behaviors} of a
distributed system are modeled by the sequence of edge labels obtained
by traversing along the edges of the graph.  Multiple edges emanating
from a single node indicate that the state represented by the node may
possibly evolve (non-deterministically in case edge labels are identical)
in different ways.

In practice, even mundane distributed systems can give rise to rather
large LTSs. Analyzing the behaviors described by such large LTSs, using
e.g., \emph{model checking},  can be quite expensive computationally.
This can sometimes be overcome by minimizing the LTS. Of course,
the behaviors, relevant to the analysis, should be preserved. In
practice, one can safely use minimization with respect to \emph{strong
bisimulation} equivalence~\cite{Par:81}; this equivalence allows one to
remove duplications of behaviors and ``fold'' unfolded behaviors.

The language mCRL2~\cite{FormalLanguagemCRL2}, which has its roots in
process algebraic theories, can be understood as a language for specifying
LTSs and reasoning about them modulo strong bisimulation. 
\emph{Processes}, representing LTSs, can be composed using operators
such as sequential composition and alternative composition to obtain new
processes. The basic building blocks of the language are \emph{actions},
representing atomic events such as \begin{math}read\end{math} and
\begin{math}send\end{math}. Special
actions (constants) are defined, such as the \emph{silent step} (${\tau}$) for
denoting internal unobservable behavior, and the \emph{deadlock} (${\delta}$) action
upon which the system cannot perform any other action.  If processes
\begin{math}p\end{math} and \begin{math}q\end{math} represent some system,
their alternative composition, denoted \begin{math}p+q\end{math},
represents the system that behaves as \begin{math}p\end{math}
when the first action executed comes from \begin{math}p\end{math},
and it behaves as process \begin{math}q\end{math} otherwise; this
models a \emph{nondeterministic choice} between behaviors. In terms
of the underlying LTS model, this may be understood as the operation
that constructs a new LTS from the LTSs of \begin{math}p\end{math}
and \begin{math}q\end{math} by creating a new top node that inherits
all edges from the top nodes of the LTSs of \begin{math}p\end{math}
and \begin{math}q\end{math}.  Sequential composition simply allows to
``glue'' the behavior of two processes. That is, \begin{math}p.q\end{math}
behaves just as process \begin{math}p\end{math}, and, upon successful
termination, it continues to behave as process \begin{math}q\end{math}.
In the underlying LTS models, this essentially creates a new graph that merges the
top node of \begin{math}q\end{math}'s LTS with the
nodes marked as terminal \begin{math}p\end{math}'s LTS.
In addition, new processes can be specified through recursion,
effectively defining LTSs that have loops, by filtering actions, by
renaming (concurrently executed) actions, and by composing processes in
parallel, etc.

Apart from describing processes, mCRL2 has a data language that
facilitates describing realistic systems where data influences the
system behavior.  The data language has built-in support for standard
data types such as Booleans (\begin{math}Bool\end{math}), and integer
(\begin{math}Int\end{math}), natural (\begin{math}Nat\end{math})
and positive (\begin{math}Pos\end{math}) numbers, and infinite lists
(\begin{math}List(\_)\end{math}), sets (\begin{math}Set(\_)\end{math})
and bags (\begin{math}Bag(\_)\end{math}) over arbitrary data types. In
addition, users have various ways in which they can define their own
data types and operations on these. For instance, one can define one's
own enumerated data types and reason about these.

Processes can be parametrized by data parameters and values. This way, the
process behavior can be influenced by the values such parameters carry;
mCRL2 offers two operators for this. First, given a process description
\begin{math}p\end{math} in which a variable $d$ occurs unbound, process
$\sum_{d:D}~p$ describes the process in which non-deterministically
a value is chosen for variable $d$. The $\sum$-operator can best be
understood as a generalization of the binary $+$ operator. Second, the
\emph{if-then-else} constructs of the form \begin{math}b\rightarrow
p\diamond q\end{math} can be used to describe the process that
behaves as process \begin{math}p\end{math} if the Boolean expression
\begin{math}b\end{math} holds and as process \begin{math}q\end{math}
otherwise.  Actions can carry data values, too, allowing one to model
the exchange of information.

Concurrent processes can interact by enforcing synchronous communication.
If actions \begin{math}send\end{math} and \begin{math}read\end{math}
are intended to happen at the same time, rather than as single isolated
actions in arbitrary order, this should be specified using the mCRL2 operator
\begin{math}comm\end{math} ($ \Gamma$) in the following manner:
$ \Gamma_{\{send|
   read} $
  $  _{\rightarrow commAction\}} $
. In practise this means that that actions \begin{math}send\end{math} and \begin{math}read\end{math}
must communicate to action \begin{math}commAction\end{math}. All occurences
of \begin{math}send | read\end{math} will be replaced by \begin{math}commAction\end{math},
provided that the data parameters that these two actions carry, match.
Such features are indispensible for modeling
distributed and concurrent systems.

\subsection{From DIRAC to mCRL2}

Any formal analysis uses a simplified description of the real system. Even in the 
best possible scenario, where the target implementation language and the modeling 
notation are very close (\cite{Java_PathFinder},\cite{Musuvathi04modelchecking}), 
it is practically impossible to avoid the use of abstraction to create a simplified model. 
Software implementations are large and often contain many details that are irrelevant 
for the intended analysis. In that sense, a model is an abstraction from reality,
where certain implementation details are ignored.

Abstraction aims at reducing
the program's state space in order to overcome the resource limitations \cite{Pelánek08fightingstate} encountered during model-checking. 
Furthermore, specification languages describe \textit{what} is being done, 
abstracting away from the details of \textit{how} things are done. 
Then, the ultimate question is: \textit{how do we establish correspondence between model and code?}

High-level design documents are a good starting point, but insufficient for 
building a sound model.
In absence of more detailed up-to-date behavioral models, 
we based our models on the source code and discussions with developers.
A popular abstraction technique
for identifying code subsets that potentially affect particular 
variables of interest (slicing criteria) is program slicing \cite{Hatcliff99slicingsoftware}. 
It reduces the behavior of a program
by removing control statements and data structures deemed irrelevant for 
the criteria. In the context of building a formal model, the slicing criteria 
depends on the focus of the final analysis, i.e., the intended properties to be checked.
The result of such an approach is typically an over-approximation of the original program behavior.
If the abstract model is checked and found to be in conformance with the property,
one can be sure that the behavior of the original program satisfies this property too. 
In other words, the obtained abstraction is sound.

Unfortunatelly, so far the only practical
applications of automated slicing have been
on C/C++ and Java programs (\cite{frama_c,codeSurfer,Corbett00bandera}). 
Research has not yet
matured on the topic of code slicing for dynamic languages, such as Python, so no automated tools exist to aid 
in the process of obtaining the abstract model. 
Therefore, we performed the program slicing manually, relying on the Eclipse IDE for 
reverse-engineering and dependency analysis of the subsystems.
Given the recurrent invalid state transitions of entities within DIRAC, 
we considered the possible race conditions caused by multiple agents updating 
the service states to be the target of our analysis. We limited the scope
to the analysis of the following entities: \textit{Tasks} (SMS), \textit{CacheReplicas} (SMS) and \textit{Jobs} (WMS).
This determines our slicing criteria. In the following, we use the SMS as a case study
for describing the established modeling guidelines. They can be 
directly applied to other DIRAC subsystems.

\subsubsection{Control Abstractions}
As already explained, all agents repeat the same logic in every subsequent iteration: first they
read some entries of interest from the service database, then they process the cached data, and finally
they may write back or update entries, based on decisions from the processing step.
They can be naturally modeled as \textit{recursive processes}. Each process represents a distinct agent,
specifying its expected behavior when interacting with the service database.
Take as an example the code snippet
in Listing~\ref{listing2}, from the \textit{Request Preparation Agent}. 
Although much of the code is omitted for clarity, the necessary parts for
illustrating the basic idea are kept.
The first highlighted statement is the selection of all "New'' CacheReplicas
entries. What follows is retrieval of their metadata from the LFC,
external to this subsystem. Subsequently, list and dictionary manipulations
are done to group the retrieved data depending on the outcome.
Two lists of replica IDs are built before the last step: one for the problematic catalog entries,
and one for the successful sanity checks.
Finally, the last two highlighted code segments update the states 
of the corresponding CacheReplicas to "Failed'' and ``Waiting'' respectively.
\definecolor{light-gray}{gray}{0.90}
\lstset{language=Python,tabsize=2,caption={\textit{RequestPreparationAgent.py} code excerpt},
  stringstyle=\ttfamily,
  %basicstyle=\small,
  frame=tb,
  breaklines=true,
  showstringspaces=false,
  boxpos=t
 }
\begin{lstlisting}[label=listing2,escapechar=!,basicstyle=\ttfamily\fontsize{8}{7}\ \selectfont]
  def prepareNewReplicas( self ):
    !\colorbox{light-gray}{res = self.getNewReplicas()}!
    if not res['Value']:
      gLogger.info('There were no New replicas found')
      return res
    [...]
    # Obtain the replicas from the FileCatalog
    res = self.__getFileReplicas( fileSizes.keys() )
    if not res['OK']:
      return res
    failed.update( res['Value']['Failed'] )
    terminal = res['Value']['ZeroReplicas']
    fileReplicas = res['Value']['Replicas']
    [...]
    replicaMetadata = []
    for lfn, requestedSEs in replicas.items():
      lfnReplicas = fileReplicas[lfn]
      for requestedSE, replicaID in requestedSEs.items():
        if not requestedSE in lfnReplicas.keys():
          terminalReplicaIDs[replicaID] = "LFN not registered at requested SE"
          replicas[lfn].pop(requestedSE)
        else:
          replicaMetadata.append((replicaID, lfnReplicas[requestedSE], fileSizes[lfn]))

    # Update the states of the files in the database
    if terminalReplicaIDs:
      gLogger.info('%s replicas are terminally failed.' % len( terminalReplicaIDs ) )
      !\colorbox{light-gray}{res = self.stagerClient.}!
	    !\colorbox{light-gray}{updateReplicaFailure(terminalReplicaIDs)}!
    if replicaMetadata:
      gLogger.info('%s replica metadata to be updated.' % len( replicaMetadata ) )
      # Sets the Status='Waiting' of CacheReplicas records 
      #that are OK with catalogue checks
      !\colorbox{light-gray}{res = self.stagerClient.}!
	    !\colorbox{light-gray}{updateReplicaInformation( replicaMetadata )}!
    return S_OK()   
\end{lstlisting}
The logging statements, although critical for operational matters, will
not affect the entries' states, and can be translated to \textit{silent steps} in mCRL2.
Furthermore, instead of tracing back and modeling all variables on which 
the two final lists depend, we can use nondeterminism. 
It is not known upfront which branch execution will follow
for a particular replica, as it depends on external behavior (i.e., the interactions of the system with its environment).
By stubbing out the communication with the LFC and
most of the local variable manipulations 
that follow, and replacing them with a \textit{nondeterministic choice} between the two
ultimate state updates, we can include both possibilities in the model behavior,
and still preserve correctness. Of course, depending on the context, some 
variable values cannot be ignored, in which case determinism can be added
using an \textit{if-then-else} mCRL2 statement. All relevant selection and update 
statements are translated into \textit{actions parametrized with data}.

\subsubsection{Data Abstractions}
The CacheReplicas entity contains more information besides the status of a file.
Every database entry has a unique identifier, descriptive data such as 
the storage where it resides, its full path, checksum, timestamps, etc., as can be seen
from the table description in Fig.~\ref{fig:CacheReplicasTable}.
\begin{figure}[bp!]
\begin{center}
 \includegraphics[width=0.5\linewidth,keepaspectratio=true]{./graphics/Figure5.png}
\end{center}
\caption{CacheReplicas table description}
\label{fig:CacheReplicasTable}
\end{figure}
Model checking can only be performed on closed models, where the domains
of all variables are finite. Since we are only interested in state transitions in terms of 
changes to the \textit{Status} field,
we can collapse most of this descriptive data, and represent this entity as a 
\textit{user-defined sort} (type) in mCRL2:
\begin{displaymath}
\begin{aligned}
sort\ CacheReplicas = \textbf{struct}\ Start\ | 
			  New\ |  \\
		    Waiting\ | 
          StageSubmitted\ | \\
		      Staged\ | 
		      Failed\ | 
		  Deleted\ ;
\end{aligned}
\end{displaymath}
This defines an enumerated data type with all possible states.
The Tasks entity is modeled in the same manner. Lists of these sorts can be easily modeled 
in mCRL2 as \begin{math}List(Tasks) \end{math} and 
\begin{math}List(CacheReplicas) \end{math}.
To define the many-to-many relationship between Tasks and CacheReplicas,
we join these data elements in a tuple:
%\begin{lstlisting}[tabsize=10,basicstyle=\ttfamily\fontsize{8}{7}\selectfont]
\begin{displaymath}sort \ Tuple = \textbf{struct}\ p(t:Nat,r:Nat,link:Bool)
\end{displaymath}
%\end{lstlisting}
The first two elements are the list positions of the Tasks and CacheReplicas
entries, while the last one indicates whether a relation between
them exists at a given moment of the system execution.

In reality agents operate on lists of IDs corresponding to the database
entries, so functions for transforming items of type \begin{math}List(CacheReplicas)\end{math}
and \begin{math}List(Tasks)\end{math}
to a list of identifiers (i.e., positions) 
\begin{math}List(Nat)\end{math} are necessary.
One such \textit{functional data transformation} used in the model is the 
$t2id:List(Tasks)\times Tasks\mapsto List(Nat)$ function, which,
given an existing list of Tasks and a specific Tasks value, returns the list positions
 matching the value.
For example:
% example t2id([one,two,two,three],two)->[1,2] 
\begin{displaymath}
t2id([Staged,New,Staged,Failed],Staged)\rightarrow [0,2] 
\end{displaymath}
Another example is the 
$id2cr:List(Nat)\times List(CacheReplicas)\times CacheReplicas\mapsto List(CacheReplicas)$                                                                                                
function, which can be used to update certain CacheReplicas list entries with a new value, in the following way:
\begin{displaymath}
id2cr([0,1],[Waiting,Staged,New],Failed) \\
 \rightarrow \\
\end{displaymath}
\begin{displaymath}
[Failed,Failed,New]
\end{displaymath}
These data transformations provide a natural way of modeling
the actual database operations.

The mCRL2 language does not support global variables (or a similar construct).
Therefore, the shared database is modeled as a wrapper process
that keeps the list of data entries as a parameter in its local memory. This
recursive process continuously listens and responds
to requests from other processes (agents), as illustrated below.

$_{proc\ CacheReplicaMem(d:List(CacheReplicas)) = }$

$_{\sum_{t:CacheReplicas} RPAgent\_selectCacheReplicas\_(cr2id(d,t),t). } $
$_{CacheReplicaMem(d) +}$

$_{\sum_{l:List(Nat),t:CacheReplicas} RPAgent\_prepareNewReplicas\_(l,t).}$
$_{CacheReplicaMem(id2cr(l,d,t))+}$

$_{\dots\ +}$

$_{\sum_{l:List(Nat),t:CacheReplicas} RFAgent\_removeReplicas\_(l,t).}$
$_{CacheReplicaMem(id2cr(l,d,t))+}$

$_{CacheReplicaMem(d);}$
$ \\ $

The information exchange is arranged via actions of the
respective processes that are enforced to communicate.
The summation operator allows the process to 
accept any value of the \begin{math}CacheReplicas\end{math}
and \begin{math}List(Nat)\end{math} sort passed by agents.
To ensure that such synchronous communication between the memory process and the agents
is possible, the mCRL2 \begin{math}allow\end{math} ($ \nabla$)
and \begin{math}comm\end{math} ($ \Gamma$)
constructs are used:

$ \nabla_{\{RPAgent\_selectCacheReplicas,
    RPAgent\_prepareNewReplicas, } $
  $    _{\dots\\, RFAgent\_removeReplicas\}} $

$ \Gamma_{\{\_RPAgent\_selectCacheReplicas|
   RPAgent\_selectCacheReplicas\_} $
  $  _{\rightarrow RPAgent\_selectCacheReplicas, } $
  $ _{\_RPAgent\_prepareNewReplicas|RPAgent\_prepareNewReplicas\_} $
  $  _{\rightarrow RPAgent\_prepareNewReplicas,\dots\\, } $
 $ _{\_RFAgent\_removeReplicas|RFAgent\_removeReplicas\_} $
  $  _{\rightarrow RFAgent\_removeReplicas\}} $
$ \\ $

To complete the picture, we give an outline of the model for the \begin{math}Request Preparation Agent\end{math} process:

$ _{RPAgent = }$
	$ _{\sum_{cc:List(Nat)} \_RPAgent\_selectCacheReplicas(cc,New). }$

	$ _{((cc\neq[])\rightarrow \ (\_RPAgent\_prepareNewReplicas(cc,Failed). } $

	$_{	\sum _{tt:List(Nat)} \_RPAgent\_selectTaskReplicas(tt,cc).  }$

	$_{	((tt\neq[])\rightarrow \_RPAgent\_prepareNewReplicasT(tt,tFailed)\ \diamond \ \tau)\ + }$

	$ _{\_RPAgent\_prepareNewReplicas(cc,Waiting))\ \diamond \ \tau) } $

$ _{.RPAgent; }$
$ \\ $

%Since the services are the frontend for all interractions with the storage,
%they essentially act as wrappers around the storage. By already modeling
%the storage as a recursive process with memory, there is no need to model them explicitely. 
As can be seen, the translated process is a sequence of
different actions for communicating with the CacheReplicas memory process. 
These include obtaining and modifying the data that this memory process maintains. 
Finally, the model is put together as a parallel composition of all
communicating processes, in the \begin{math}init\end{math}
part of the specification.

$ _{init \ RPAgent\ ||\ SRAgent\ ||\ \dots\ ||\ CacheReplicaMem([]);}$
$ \\ $

Although the WMS model is larger, it is constructed using the same approach.
The complete models are available at \cite{svn_mcrl2}.
\section{Analysis and Issues}
\label{sec:Section_4}
The purpose of model checking is to prove that the modeled system
exhibits certain behavior (requirement), or alternatively, to discover problems. 
The operation of a model checker closely resembles graph traversal.
Nodes of the graph represent the states of the system, while the edges
connecting them represent transitions, or state changes. The collection of 
all possible states and transitions forms the state space, or an LTS.
Typically, model checkers
examine all reachable states and execution paths in a systematic and 
fully automated manner, to check if a certain property holds.
In case of violation of the examined property, a counterexample is 
provided as a precise trace in the model, showing which interleavings
of actions of the parallel components led to the violation.
After the model is written, the state space can be 
explicitly generated and stored. 
The execution times and the resulting numbers of states are
presented in Table~\ref{mcrl2Stats}. Generating the LTS with the mCRL2 toolset can be a time consuming process,
since the state space typically grows exponentially with the number of parallel processes in the model.
\begin{table}[h!]
\vspace{+4 pt}
  \centering
  \begin{tabular}[1.0\textwidth]{  p{1cm} | l  p{1cm} p{1cm}  p{1.6cm}  }
    \hline
  \rowcolor[gray]{0.9}
   System & States & mCRL2 LoC & Python LoC & Generation time \\ \hline\hline
   SMS & 18,417 & 432 & 2,560 & \textless 10 sec. \\ \hline
   WMS & 160,148,696 & 663 & 15,042 & $\sim$12 hr. \\
    \hline
  \end{tabular}
  \caption{mCRL2 model statistics}
\label{mcrl2Stats}
\end{table}
\subsection{Simulation and debugging}
Apart from verification, the mCRL2 toolset has a rich set of tools for analysis of the modeled system. 
The XSim simulator allows to replay scenarios and inspect in 
detail the current state and all possible transitions in the model, for every execution step. 
Modeling errors are not uncommon, especially in the first(draft) versions of the
model, and XSim provides the necessary feedback, acting as a debugger.
This process was already valuable for understanding the
system and identifying interesting behaviors that were later included
in the requirements. We want to stress that this is especially useful
when building models of existing implementations, where at first glance it 
is not very clear which (un)desired properties need to be formulated and automatically probed,
before they actually surface in the real system.
\begin{figure*}[bp!]
  \centering
  \subfloat[Logging info of a DIRAC job]{\label{fig:Job-Logging-Info}\includegraphics[width=0.4566\linewidth]{./graphics/Figure6a.png}}    
  \hfill
  \subfloat[XSim simulator trace]{\label{fig:simulator}\includegraphics[width =0.5\linewidth]{./graphics/Figure6b.png}}
  \caption{Invalid job state transitions}
  \label{fig:Simulator-job-trace}
\vspace{-8 pt}
\end{figure*}
One such problematic behavior was recently reported in DIRAC's production (Fig.~\ref{fig:Job-Logging-Info}), where
a job had transited between two different terminating states, ``Failed'' and ``Done''.
Replaying the behavior with XSim revealed the exact same trace in the WMS model (Fig.~\ref{fig:simulator}).
This was a result of several (difficult to reproduce in reality) circumstances:
The \textit{JobWrapper} process was continuously trying to access a file already 
migrated to tape. Meanwhile, the \textit{Stalled Job Agent} responsible for 
monitoring the pilot declared the job as ``Stalled'', and ultimately as non-responsive (``Failed''),
since its child process was busy with data access attempt for a long time.
However, once data access succeeded, the \textit{JobWrapper} finally started executing the 
actual payload and reported a ``Running'' status. 
The \textit{JobWrapper} assumes that nothing has happened to the status of a job once 
it brings it to a ``Running'' state, 
and only reports a different MinorStatus value afterwards.
The logic is such
because it is naturally expected that this process has exclusive control 
over the rest of the job workflow, once it starts running. Fixing such
a problem without compromising efficiency is not trivial.
\subsection{Visualization}
\begin{figure}[tp]
\vspace{-10 pt}
\includegraphics[width=0.5\linewidth,keepaspectratio=true]{./graphics/Figure7_new.png}
\centering
\caption{CacheReplicaMem process visualization with DiaGraphica}
\label{fig:DiaGraphica}
\end{figure}%123478
Reasonably small LTSs can be easily visualized with the interactive GUI tools,
as the tools employ smart clustering techniques to reduce the complexity of the image.
For instance, the SMS model, with around 18000 states,
was visualized with DiaGraphica. Fig.~\ref{fig:DiaGraphica} shows a projection (clustering) of the state space
on the CacheReplicas memory process, which resulted in only 6 interesting states for this process alone.

The clustering
allowed us to see precisely which chain of actions by the concurrent agents
advanced the state of the CacheReplicas entity. In fact, in the modeling phase we discovered the first problematic
SMS behavior, blocking the progress of staging
tasks (and as a result the progress of jobs). The origin of the deadlock had been a
trivial human logic flaw introduced in coding. At a particular circumstance,
when a ``New'' task enters the system \textbf{and} all associated replicas
are already ``Staged'' (possibly by other jobs requiring the same input files),
its status would immediately be promoted to ``Done''.
Such tasks would never be picked up by SMS and called-back to the appropriate job, 
since the agent responsible for this final step would only look for ``Staged'' tasks.
A proper state machine implementation
instead of hardcoding states in SQL queries can prevent such bugs.
\begin{figure}[bp!]
\vspace{-10 pt}
\includegraphics[width=0.57\linewidth,keepaspectratio=true]{./graphics/Figure8.png}
\centering
\caption{State-space visualisation of the SMS with ltsview}
\label{fig:LTSviewVisualisation}
\end{figure}%123478

The ltsview tool can be used for a more sophisticated way of visualizing model state spaces in 3D.
It also employs clustering of states based on structural properties, but serves to further 
explore symmetry in the behavior of the system and discover unexpected visual anomalies.
The SMS state space is shown in Fig.~\ref{fig:LTSviewVisualisation}. 
The behavior 
starts at the top of the cluster and proceeds downwards. Branching points indicate 
alternatives in the behavior of the system. It is possible to mark and color different
regions based on values of the state vectors, as well as transitions based on their labels.
The particular figure shows a suspicious transition between ``Failed'' and ``Staged'' tasks presented
by distinct coloring of the regions. One can also interactively simulate the behavior while keeping 
track of the previous, current state, and the possible transitions.

\subsection{Model checking}
The default requirement specification language in the mCRL2 toolset is
the modal $\mu$-calculus \cite{ProcessesWithData}, extended with regular
expressions and data.  For our purposes, the regular expressions suffice.
These typically allow for specifying that certain behaviors must be
absent or present, and \emph{if-then} scenarios. Regular expressions
are constructed using constants $true$ and $false$ and the modalities
``[]'' and ``\textless\textgreater''. The constants $true$ and $false$
have their usual meaning: $true$ holds in every state, while $false$ does
not hold in any state of the model. The modalities are used for expressing
\textit{necessity} (``[]'') and \textit{possibly} (``\textless\textgreater'')
statements in a formula. \textit{Necessity} means that the formula should
hold for every behavior within the brackets, while \textit{possibly}
means that there exists a behavior where the formula is satisfied. The
behaviors inside the modalities can be specified concisely using
\emph{action formulae}. These can be constructed using actions and the
operations ``$\mid\mid$'' for union and ``!''
for the set complement; the action formula \begin{math}true\end{math}
represents the set of \emph{all} actions. Any action or action formula
can be turned into a set of behaviors using the ``$\ast$'' operator,
which is used for expressing cardinality, meaning that actions can occur
any number of times; the ``.'' operator allows for concatenating
behaviors. In addition to these constructs, one can use
quantification over data variables, Boolean expressions,
etc. In mCRL2, the model checking problem is often converted to
an equation system solving problem, see~\cite{ProcessesWithData}; such equation systems
can be manipulated and solved, thereby answering the encoded
model checking problem.

Using the gained understanding of the system behavior, as well as the
reported and anticipated problems, we formulated several requirements for
the model checker to systematically probe, see below. Typical expressions
are of the form ``[$A$]$false$'', which holds when a system has no
behaviors matching $A$, and ``[$A$]\textless$B$\textgreater$true$'',
which holds if a system has a behavior matching $B$ whenever a behavior
matching $A$ has occurred.

\lstset{tabsize=1,
  stringstyle=\ttfamily,
  caption={},
  %basicstyle=\small,
  breaklines=true,
  showstringspaces=false,
  morekeywords={tStaged,tDeleted,tFailed,tStageSubmitted,tNew,Done,Killed,MarkedForTermination,Running,JobInitialization}
 }
\renewcommand{\labelenumi}{\arabic{enumi}.}

\begin{enumerate}
\item Each task in a terminating state (``Failed'' or ``Staged'') is eventually removed from the system. (\textit{progress})

\begin{lstlisting}[basicstyle=\ttfamily\fontsize{8}{9}\selectfont]
[true* .(state([tStaged]) || state([tFailed])).(!(state([tDeleted])))*] 
<(!(state([tDeleted]))* .(state([tDeleted]))> true
\end{lstlisting}

\item A deleted task will never be referenced for transition to any other state. (\textit{safety})
\begin{lstlisting}[basicstyle=\ttfamily\fontsize{8}{9}\selectfont]
[true*.state([tDeleted]). true*. 
(state([tNew]) || state([tStageSubmitted]) || state([tStaged]) || state([tFailed]))]false
\end{lstlisting}
\item Once a job has been killed, it cannot ressurect and start running. (\textit{safety})

\begin{lstlisting}[basicstyle=\ttfamily\fontsize{8}{9}\selectfont]
[true*.state([jobstatus(Killed,MarkedForTermination)]).true*.state([jobstatus(Running,JobInitialization)])]false
\end{lstlisting}

\end{enumerate}

\begin{figure}[tp]
\includegraphics[width=0.7\linewidth,keepaspectratio=true]{./graphics/Figure9.png}
%was PropertyViolation1.png
\centering
\caption{Violation of requirements 1 (top) and 2 (bottom)}
\label{fig:PropertyViolation}
\end{figure}
Model checking was carried out on a 64bit Intel Core\texttrademark2 Duo (1.6GHz) machine with 2GB RAM. 
Both requirements 1 and 2 were found to be violated, as can be seen from the traces (Fig.~\ref{fig:PropertyViolation}).
The race conditions manifest themselves in a few subtle interactions between the SMS storage and the agents.
In both cases, at step 4 the \textit{Stage Request Agent} issues prestage requests and 
as a result the CacheReplicas entries (second element of the State column) are updated to ``StageSubmitted''. Between the moment
that this agent has selected the corresponding Task to update to the same state,
to the point where the update is done (last step in both traces),
other agents may have monitored these replica records and updated
them to a new state, along with the associated Task. In practice, the manifestation
of such race conditions depends on the speed of the agents propagating the
state changes between the selection and update done by the \textit{Stage Request Agent}.
Such cases are nevertheless encountered in reality, when this agent
has large lists of records to process. This results in a deadlock situation, as 
the Task will have no further state updates made by other agents, since
from their perspective the state propagation is finished.

Model checking and debugging with explicit state space generation was not
a viable option for the WMS, due to its size (13 concurrent processes
giving rise to a state space of over 150 million states). We therefore
resorted to using the LTSmin symbolic reachability tool~\cite{BP:08} and
the symbolic equation system solver built on top of LTSmin's equation
system explorer~\cite{KP:12}.  These tools are almost instantaneous
in traversing the state space and equation systems, respectively,
taking less than 20 seconds. They rely on the use of clever data
structures such as BDDs~\cite{BCMDH:90} for concisely representing the
underlying LTS or equation system that is being explored. Even though
all information for computing the explicit artifacts is present in such
data structures,  obtaining these from such data structures can again
be costly, computationally.

The symbolic equation system explorer and solver effectively solves the
model checking problems for our requirements. However, in case of violations
of the requirements, the symbolic verification tools currently do not give
counterexamples. In order to understand the violations, we employ a
different trick. We use the symbolic state space exploration tool's
option  
for tracking the occurrence of a certain (specified) action and
reporting a trace if such an action is encountered during exploration. 
For this purpose, we extended the specification 
with a monitoring process which fires an \textit{error} action if requirement 3 is
violated. This monitoring process is set to run in parallel with the system, and
observes the relevant actions and state changes that the system itself takes.

A generic monitoring process for checking temporal properties of the form:
\begin{lstlisting}[basicstyle=\ttfamily\fontsize{8}{9}\selectfont]
[true*. state(l1). true*. state(l2)]false
\end{lstlisting}
is shown below.

$ _{proc\ Monitor(l1,l2:List(Job), b:Bool) = }$

	$ _{\tau .Monitor(l1,l2,b) }$

	$ _{+\ (!b\ \rightarrow \ (\sum_{l:List(Job)} . (l == l1)\ \rightarrow \ (monitor(l). Monitor(l1,l2,true)) }$

	$ _{ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \diamond \ (monitor(l). Monitor(l1,l2,b) )) }$

	$_{\ \ \ \ \ \ \diamond \ (\sum_{l:List(Job)} . (l == l2)\ \rightarrow \ (monitor(l). error(l1,l2)) }$

	$ _{ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \diamond \ (monitor(l). Monitor(l1,l2,b) ) }$

	$ _{\ \ \ );}$
$ \\ $

The automaton of the concrete monitor instance which accepts requirement 3 trace is shown in Fig.~\ref{fig:monitorProcess}. 
\begin{figure}[tp]
\includegraphics[width=0.5\linewidth,keepaspectratio=true]{./graphics/Figure10.png}
\centering
\caption{Monitor automaton for requirement 3}
\label{fig:monitorProcess}
\end{figure}%123478
Whenever the observed system changes the state of a job to ``Killed'', the automaton also changes
its state to L1 by enforcing synchronous communication between the respective actions.
The silent step (${\tau}$) can be executed as a result of any non-relevant system state change
from the perspective of the automaton. As soon as the system changes the state of the job further
to ``Running'', the automaton makes a move in a similar fashion, after which point the action 
\textit{error} is executed, and can be detected by the LTSmin tool. 
\begin{figure}[bp]
\includegraphics[width=0.6\linewidth,keepaspectratio=true]{./graphics/Figure11.png}
\centering
\caption{"Zombie" job starts running after being killed}
\label{fig:zombieJob}
\end{figure}%123478
If such a full cycle is accepted by the automaton, the temporal property is indeed violated.
Using this technique, a counterexample (Fig.~\ref{fig:zombieJob}) showed that, when 
staging was involved in the workflow of a job, the callback from
the SMS was not properly handled.
It awakened the job immediately to a ``Checking'' (and eventually ``Running'') state
even if it had been manually ``Killed'' 
(using DIRAC's command line API) by production managers meanwhile. Such zombie jobs
were discovered in the system on several occasions,
and with the model at hand it was easy to replay the behavior and localize the
problem. The implementation was fixed to properly guard the callback against this particular case.
The changes were also reflected back to the mCRL2 model ~\cite{svn_mcrl2}. Fig.~\ref{fig:zombieJobFix}
shows that the stager callback no longer affects the job state if it has been changed from its
original ``Staging'' one in the meantime. 
The state space of the fixed model amounts to 
157048664 states, the reduction being a result of the blocking of the unwanted behavior.
However, we must note that model checking still discovered traces where the property 
is violated, even without the stager being involved.
The current architecture does not permit an easy workaround for this problem.
Protecting shared memory access using traditional locking techniques does not scale well.
There is an ongoing effort in redesigning DIRAC's architecture, to accomodate
a task dispatcher that takes care of persisting the state of the tasks to the storage 
backend and distributing them among so called ``executors''.This should 
replace the old agent components' polling architecture, and reduce the load
they place on the storage, in addition to eliminating race conditions.

%TODO: fix the model, show a fixed trace
\begin{figure}[tp]
\includegraphics[width=0.6\linewidth,keepaspectratio=true]{./graphics/Figure12.png}
\centering
\caption{ Race condition involving the stager callback no longer occurs}
\label{fig:zombieJobFix}
\end{figure}%123478

\section{Conclusions}
\label{sec:Section_5}
In this paper we have applied a formal methods approach for analyzing CERN's 
DIRAC grid system. 
We used two components as case studies: the Workload
Management and the Storage Management System, which are the driving force 
of DIRAC.
By creating an abstract model, simulating, visualizing, and model checking it with the mCRL2 toolset,
we were able to gain insight into the system behavior and detect critical
race conditions and livelocks, which  
were confirmed to occur in the real system. 
Despite the continuous DIRAC development for 10 years, these problems 
were notoriously difficult to discover and trace, considering the number of concurrent components involved. 
Testing of concurrent systems and reproducing error traces is often challenging because of the lack of 
control that the tester has over the execution of the concurrent processes. 
With the model at hand, replaying
the traces and localizing the problems became much more straightforward, overcoming many limitations of
traditional testing techniques for large-scale distributed systems.

The investment in writing an abstract model in a formal language 
pays off proportionally to the size (number of components) of the distributed system. 
This holds especially for grid systems where a large number of concurrent
 components of the same kind deliver the overall system functionality. 
Once becoming proficient with
the formal language notation, modeling such components is
reduced to reapplying similar abstraction rules.
Practitioners
of model checking have already built sufficiently mature tools
that can be utilized (almost) as blackboxes by more traditional
software developers, hiding the mathematical theories under
the hood. Although some domain-specific knowledge is always necessary,
the descriptive guidelines we have proposed for obtaining
an abstract model can be easily 
applied to other large-scale industrial systems of a similar nature.
It goes without saying that the analyst must be careful to faithfully
describe the system behavior in the abstract model. With this critical step in mind, formal methods
are an important addition to the assembly of software analysis and verification techniques for concurrent distributed systems.
\section{Future work}
\label{sec:Section_6}
Besides discovering behavioral problems, the 
model gives opportunities to analyze the system performance in the future.
Performance analysis aims at predicting the behavior of a system with respect to dynamic properties, such as 
the expected throughput, response time, resource usage, number of requests per unit time, etc.
For this, the formal model must be enriched with timing and probability quantities attached 
to certain activities of interest.  Adding this kind of information to traditional process algebra models
means getting in the domain of stochastic process algebra. Random variables with exponential distribution are
typically associated with the designated actions, to express system transitions with duration, which could be estimates
of processing time or communication overhead. Tools based on stochastic process algebra formalisms can
then generate the underlying Markov Chain model representation and solve it to obtain steady-state probability distributions over the state space. These measures can provide a good base for assessing the system's efficiency and reliability. 
Furthermore, by varying the quantitative inputs, alternative designs can be evaluated before actual implementation
takes place. One such stochastic process algebra is PEPA (Performance Evaluation Process Algebra \cite{PEPA}), supported
by an Eclipse Plug-in for modeling and analysis of systems. 
PEPA models are annotated with information about the duration of activities and their relative probabilities of occurence.

In the context of DIRAC, our future work will be focused on extending the mCRL2 models with information about polling intervals of the agents, the rate at which database queries can be executed, and the overhead due to communication with external components (such as the storage middleware and the LFC), in order to obtain estimates of the agents utilization, mean processing time before a file replica is staged, and the average cache occupancy. 
Furthermore, we want to assess an alternative DIRAC WMS architecture which is claimed to reduce the reaction time of the system with respect to the optimizers-chain part of the job workflow. We want to have an indication of the possible system bottlenecks under different load (job accumulation), by varying the number of instances of certain concurrent components. This will give us valuable information on how to proceed with deployment in production environments.

Fortunately, we do not have to use another formalism for enhancing the models with stochastic information. The CADP toolset \cite{CADP2010} for analysis of stochastic models is well integrated with mCRL2 for this purpose. CADP can explore and manipulate mCRL2 specifications via an interface for a common LTS format representation \cite{CADPPerformanceEval} which both tools understand. 
\section*{Acknowledgements}
The authors would like to thank Wan Fokkink from VU University Amsterdam, Dragan Bosnacki from TU Eindhoven,
as well as Ricardo Graciani, Philippe Charpentier and the rest of the LHCb collaboration for their valuable contribution to this work.
\bibliographystyle{elsarticle-num} 
\bibliography{myLibrary}

\end{document}

### 2013-02-04 graphics/ElsevierFGCS_singleColumn.tex
FGCS

--- 2013-02-04 ElsevierFGCS_singleColumn.tex
@@
 It provides a concurrent use of over 60K CPUs and 10M file replicas distributed over
-tens of grid sites, provided by the WLCG computing grid \cite{Bird:2011zz}. The community of users has grown to several hundreds, loading the 
-LHC grid resources with up to 30K simultaneously running jobs (Fig.~\ref{fig:LHCb-resource-usage}, top) and over 100000 jobs executed per day, during peak processing periods.
-About 10PB of data is stored and visible in the LHCb file catalogs, with daily data transfers among storage units reaching 3Gbps (Fig.~\ref{fig:LHCb-resource-usage}, bottom).
-\begin{figure*}[tp!]
-  \centering
-  {\label{fig:Job-throughput}\includegraphics[width=0.7\linewidth]{./graphics/Figure1_1.png}}    
-  \vfill
-  \vspace{14pt}
-{\label{fig:Data-transfers}\includegraphics[width =0.7\linewidth]{./graphics/Figure1_2.png}}
-  \caption{LHCb resource usage}
-  \label{fig:LHCb-resource-usage}
-\end{figure*}
+tens of grid sites. The community of users has grown to several hundreds, loading the 
+LHC grid resources with up to 50K simultaneously running jobs during peak processing periods.
 Python was chosen as the implementation language, since it enables
@@
 \begin{figure}[tp]
-\includegraphics[width=0.5\textwidth] {./graphics/Figure1.png}
+\includegraphics[width=0.5\textwidth] {./Figure1.png}
 \centering
@@
 \end{figure}
-
 The \textit{services} are passive components that react to requests from their clients,
@@
 for access. The physical resources are controlled by the site managers and made
-available through middleware services such as gLite \cite{Laure06programmingthe}.
+available through middleware services such as gLite \cite{gLite}.
 
@@
 
-The DIRAC hardware configuration of LHCb is still by far the most powerful one: it consists of about 10 servers (VO-boxes) located at CERN, hosting the central
-DIRAC services, including 6 MySQL database servers. Small fraction of the servers and databases are used for testing purposes, and are connected
-to the same pool of WLCG computing and storage resources. Each of the six LHCb Tier-1 computing centers accross Europe hosts an additional server as part of this cluster for redundancy,
-to ensure high service availability at any moment. Several other experiments are currently using DIRAC on an every day basis, 
-with new user groups evaluating its functionality and performance before adopting it in production.
-
 In the following, we focus on two related subsystems that are considered the backbone of DIRAC. These
@@
 \begin{figure}[bp]
-\includegraphics[width=0.7\textwidth]{./graphics/Figure2.png}
+\includegraphics[width=0.7\textwidth]{./Figure2.png}
 \centering
@@
 quotas, which are decided within the LHCb community, without the need of any
-effort from the grid sites.  Originally, the DIRAC WMS submitted pilot jobs using 
-the native gLite middleware tools, but this method was shown to be suboptimal. The gLite resource brokers
-are centralized components, and as such they do not have the capacity and flexibility to cope with the load of 
-a community such as LHCb. Using multiple resource brokers, rather than a single one, was not the solution:
-all brokers use the same site state information and would always submit jobs to the most suitable one globally, causing overloads to
-certain sites while underloading others very often. In addition, the gLite information system was not reactive enough to propagate
-changes in site status information to the brokers. With the wide deployment of the CREAM CE service \cite{Aiftimiei2010654}, DIRAC's WMS switched
-to submitting pilot jobs directly to sites. By interrogating the Computing Element status directly, the system reactiveness
-improved dramatically, making sites compete with each other for jobs and have a faster job turnaround.
+effort from the grid sites.
 
@@
 \begin{figure}[tp]
-\includegraphics[width=0.7\linewidth,keepaspectratio=true]{./graphics/Figure3.png}
+\includegraphics[width=0.7\linewidth,keepaspectratio=true]{./Figure3.png}
 \centering
@@
 \begin{center}
- \includegraphics[width=0.5\linewidth,keepaspectratio=true]{./graphics/Figure4.png}
+ \includegraphics[width=0.5\linewidth,keepaspectratio=true]{./Figure4.png}
 \end{center}
@@
 Apart from describing processes, mCRL2 has a data language that
-facilitates describing realistic systems where data influences the
+facilitate describing realistic systems where data influences the
 system behavior.  The data language has built-in support for standard
@@
 the criteria. In the context of building a formal model, the slicing criteria 
-depends on the focus of the final analysis, i.e., the intended properties to be checked.
+depends on the focus of the final analysis, i.e. the intended properties to be checked.
 The result of such an approach is typically an over-approximation of the original program behavior.
@@
 It is not known upfront which branch execution will follow
-for a particular replica, as it depends on external behavior (i.e., the interactions of the system with its environment).
+for a particular replica, as it depends on external behavior (i.e. the interactions of the system with its environment).
 By stubbing out the communication with the LFC and
@@
 \begin{center}
- \includegraphics[width=0.5\linewidth,keepaspectratio=true]{./graphics/Figure5.png}
+ \includegraphics[width=0.5\linewidth,keepaspectratio=true]{./Figure5.png}
 \end{center}
@@
   $  _{\rightarrow RPAgent\_selectCacheReplicas, } $
+$\\$
   $ _{\_RPAgent\_prepareNewReplicas|RPAgent\_prepareNewReplicas\_} $
   $  _{\rightarrow RPAgent\_prepareNewReplicas,\dots\\, } $
+$\\$
  $ _{\_RFAgent\_removeReplicas|RFAgent\_removeReplicas\_} $
@@
   \centering
-  \subfloat[Logging info of a DIRAC job]{\label{fig:Job-Logging-Info}\includegraphics[width=0.4566\linewidth]{./graphics/Figure6a.png}}    
+  \subfloat[Logging info of a DIRAC job]{\label{fig:Job-Logging-Info}\includegraphics[width=0.4566\linewidth]{./Figure6a.png}}    
   \hfill
-  \subfloat[XSim simulator trace]{\label{fig:simulator}\includegraphics[width =0.5\linewidth]{./graphics/Figure6b.png}}
+  \subfloat[XSim simulator trace]{\label{fig:simulator}\includegraphics[width =0.5\linewidth]{./Figure6b.png}}
   \caption{Invalid job state transitions}
@@
 \vspace{-10 pt}
-\includegraphics[width=0.5\linewidth,keepaspectratio=true]{./graphics/Figure7_new.png}
+\includegraphics[width=0.5\linewidth,keepaspectratio=true]{./Figure7.png}
 \centering
@@
 was visualized with DiaGraphica. Fig.~\ref{fig:DiaGraphica} shows a projection (clustering) of the state space
-on the CacheReplicas memory process, which resulted in only 6 interesting states for this process alone.
+on the CacheReplicas memory process, which resulted in only 7 interesting states for this process alone.
 
@@
 \vspace{-10 pt}
-\includegraphics[width=0.57\linewidth,keepaspectratio=true]{./graphics/Figure8.png}
+\includegraphics[width=0.5\linewidth,keepaspectratio=true]{./Figure8.png}
 \centering
@@
 \begin{figure}[tp]
-\includegraphics[width=0.7\linewidth,keepaspectratio=true]{./graphics/Figure9.png}
+\includegraphics[width=0.7\linewidth,keepaspectratio=true]{./Figure9.png}
 %was PropertyViolation1.png
@@
 \begin{figure}[tp]
-\includegraphics[width=0.5\linewidth,keepaspectratio=true]{./graphics/Figure10.png}
+\includegraphics[width=0.5\linewidth,keepaspectratio=true]{./Figure10.png}
 \centering
@@
 \begin{figure}[bp]
-\includegraphics[width=0.6\linewidth,keepaspectratio=true]{./graphics/Figure11.png}
+\includegraphics[width=0.6\linewidth,keepaspectratio=true]{./Figure11.png}
 \centering
@@
 \begin{figure}[tp]
-\includegraphics[width=0.6\linewidth,keepaspectratio=true]{./graphics/Figure12.png}
+\includegraphics[width=0.6\linewidth,keepaspectratio=true]{./Figure12.png}
 \centering

### 2013-02-04 ElsevierFGCS_singleColumn_old.tex
Keep old version before review.

--- 2013-02-04 ElsevierFGCS_singleColumn.tex
@@
 certificates.
-
-The DIRAC hardware configuration of LHCb is still by far the most powerful one: it consists of about 10 servers (VO-boxes) located at CERN, hosting the central
-DIRAC services, including 6 MySQL database servers. Small fraction of the servers and databases are used for testing purposes, and are connected
-to the same pool of WLCG computing and storage resources. Each of the six LHCb Tier-1 computing centers accross Europe hosts an additional server as part of this cluster for redundancy,
-to ensure high service availability at any moment. Several other experiments are currently using DIRAC on an every day basis, 
-with new user groups evaluating its functionality and performance before adopting it in production.
 
@@
 \vspace{-10 pt}
-\includegraphics[width=0.5\linewidth,keepaspectratio=true]{./graphics/Figure7_new.png}
+\includegraphics[width=0.5\linewidth,keepaspectratio=true]{./graphics/Figure7.png}
 \centering
@@
 \vspace{-10 pt}
-\includegraphics[width=0.57\linewidth,keepaspectratio=true]{./graphics/Figure8.png}
+\includegraphics[width=0.5\linewidth,keepaspectratio=true]{./graphics/Figure8.png}
 \centering

### 2013-02-11 ElsevierFGCS_singleColumn.tex
Related work...

@@
 tens of grid sites, provided by the WLCG computing grid \cite{Bird:2011zz}. The community of users has grown to several hundreds, loading the 
-LHC grid resources with up to 30K simultaneously running jobs (Fig.~\ref{fig:LHCb-resource-usage}, top) and over 100000 jobs executed per day, during peak processing periods.
-About 10PB of data is stored and visible in the LHCb file catalogs, with daily data transfers among storage units reaching 3Gbps (Fig.~\ref{fig:LHCb-resource-usage}, bottom).
+LHC grid resources with up to 30K simultaneously running jobs (Fig.~\ref{fig:LHCb-resource-usage}, top) and over 100K jobs executed per day, during peak processing periods.
+About 10PB of data is stored and visible in the LHCb file catalogs, with daily data transfers between storage units reaching 3Gbps (Fig.~\ref{fig:LHCb-resource-usage}, bottom).
 \begin{figure*}[tp!]
@@
 table. Other auxiliary tables maintain the relationship between these entities.
-\begin{figure}[tp]
-\begin{center}
- \includegraphics[width=0.5\linewidth,keepaspectratio=true]{./graphics/Figure4.png}
-\end{center}
+
+\begin{figure}
+\centering
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=0.98\linewidth]{./graphics/Figure4.png}
   \caption{DIRAC Storage Management System}
   \label{fig:DIRAC-SMS}
+\end{minipage}%
+\begin{minipage}{.5\textwidth}
+  \centering
+  \includegraphics[width=0.95\linewidth]{./graphics/SMS_CacheReplicasSM.png}
+  \caption{CacheReplicas state machine}
+  \label{fig:CacheReplicas-SM}
+\end{minipage}
 \end{figure}
+
 The processing begins with the \textit{Request Preparation Agent}. It selects all the
@@
 there are no more associated \textit{Tasks} for particular replicas, the respective
-\textit{CacheReplicas} entries are also removed.
+\textit{CacheReplicas} entries are also removed. The state machine of the \textit{CacheReplicas}
+entity is shown in Fig.~\ref{fig:CacheReplicas-SM}.
 
@@
 
+\section{Related Work}
+\label{sec:RelatedWork}
+Model checking belongs to a class of mathematically-based techniques for specification, analysis,
+and verification of software and hardware systems. Theorem proving, static checking, run-time verification, and model-based testing
+also belong to this category of formal methods. The main premise behind their use is the expectation that
+the initial investment in becoming proficient and developing models pays off in the long run, in terms 
+of improved reliability and safety of the system. While traditional testing is almost
+exclusively used as a verification methodology in many practical industrial cases,
+formal methods have started receiving more attention in the last couple of decades. 
+Testing is often based on the intuition and experience of the tester, and, as such, does not 
+always reveal subtle system errors. Hence, particular efforts are directed at
+the practical use of formal methods in the area of critical, concurrent, and distributed 
+systems.
+The application of such methods in communication protocols~\cite{SPIN_case_study,SlidingWindowProtocol,sessionMgmtMasterThesis}, 
+safety-critical software~\cite{Java_PathFinder,CMS_LHC,Mathijssen:2006:VDA:1757571.1757584},
+driver design~\cite{Ball04slamand,Linux_driver}, among others, has been a success.
+Case studies in the domain of high-performance computing, and in particular concurrent/MPI-based program design,
+have surfaced as well~\cite{Siegel04verificationof,Matlin:2002:SPS:645881.672236,Siegel:2007:MCN:1763048.1763053}, mostly 
+relying on the the MPI extension of the popular model checker SPIN~\cite{Siegel:2007:VPP:2396095.2396105}.
+Ironically enough, the model checking tools themselves do not take advantage of distributed infrastructures, and 
+as such do not scale well.
+
+To overcome the fundamental problem of state space explosion, several
+variants to explicit model checking have gained popularity. Runtime verification~\cite{citeulike:10198274,Havelund:2008:RVC:1424465.1424470}
+aims to increase confidence in the correctness, but does not make claims for absence of defects in a system.
+The target program is instrumented manually, or automatically by scanning function calls and variable assignments, 
+in order to capture and emit relevant events which are then monitored against a special component during execution.
+This component resembles our monitoring process automaton used to cope with state space explosion during model checking 
+of DIRAC's WMS system. While runtime verification is useful for partially-capturing incorrect behavior of the system implementation, 
+code instrumentation is not always a feasible option, considering the increase of memory consumption. 
+Similar to this lightweight form of verification is static program analysis~\cite{Schmidt98programanalysis}. Rather than code instrumentation and execution,
+static program analysis aims to determine runtime properties of systems at compilation time.
+A special analyzer targeted for the implementation language maintains information on the control flow and function call graphs
+of the software. Approximation is used by such tools in practice, in order to improve scalability. The obvious advantage
+of this approach is that it works directly on the implementation. However, most static analysis tools aim at discovering 
+more general implementation deficiencies, such as uninitialized variables, null pointer references, or out of bound array indices. 
+Both runtime verification and static analysis can be seen as additions to
+traditional model checking, rather than alternatives. Some progress has been 
+made in the direction of combining the approaches~\cite{Java_PathFinder,Linux_driver,Brat01combiningstatic}. For instance, Java PathFinder~\cite{Java_PathFinder}
+relies on static analysis to reduce the state space before systematically exploring it by executing the abstracted program. In~\cite{Brat01combiningstatic}
+static analysis and model checking also operate in a feedback loop. The drawback of static analyzers is their 
+practical availability for languages other than C and Java. To the best of our knowledge, no such tool exists for Python.
+However, our approach of using code slicing to abstract the implementation shares similarities.
+
+TODO: trace mining
+
 \section{Conclusions}

### 2013-02-11 ElsevierFGCS_singleColumn.tex
Trace mining added to Related Work. Deductive verificaiton TBD.

@@
 
-To overcome the fundamental problem of state space explosion, several
-variants to explicit model checking have gained popularity. Runtime verification~\cite{citeulike:10198274,Havelund:2008:RVC:1424465.1424470}
-aims to increase confidence in the correctness, but does not make claims for absence of defects in a system.
-The target program is instrumented manually, or automatically by scanning function calls and variable assignments, 
-in order to capture and emit relevant events which are then monitored against a special component during execution.
-This component resembles our monitoring process automaton used to cope with state space explosion during model checking 
-of DIRAC's WMS system. While runtime verification is useful for partially-capturing incorrect behavior of the system implementation, 
-code instrumentation is not always a feasible option, considering the increase of memory consumption. 
-Similar to this lightweight form of verification is static program analysis~\cite{Schmidt98programanalysis}. Rather than code instrumentation and execution,
-static program analysis aims to determine runtime properties of systems at compilation time.
-A special analyzer targeted for the implementation language maintains information on the control flow and function call graphs
-of the software. Approximation is used by such tools in practice, in order to improve scalability. The obvious advantage
-of this approach is that it works directly on the implementation. However, most static analysis tools aim at discovering 
-more general implementation deficiencies, such as uninitialized variables, null pointer references, or out of bound array indices. 
+To overcome the problem of state space explosion, several
+variants to explicit model checking have gained popularity. Runtime
+verification~\cite{citeulike:10198274,Havelund:2008:RVC:1424465.1424470}
+aims to increase confidence in the correctness, but does not make claims for
+absence of defects in a system.
+The target program is instrumented manually, or automatically by scanning
+function calls and variable assignments, 
+in order to capture and emit relevant events which are then monitored against a
+special component during execution.
+This component resembles our monitoring process automaton used to cope with
+state space explosion during model checking 
+of DIRAC's WMS system. While runtime verification is useful for
+partially-capturing incorrect behavior of the system implementation, 
+code instrumentation is not always a feasible option, considering the increase
+of memory consumption.
+
+Similar to this lightweight form of verification is static program
+analysis~\cite{Schmidt98programanalysis}. Rather than code instrumentation and
+execution, the objective of 
+static program analysis is to determine runtime properties of systems at
+compilation time.
+A special analyzer targeted for the implementation language maintains
+information on the control flow and function call graphs
+of the software. Approximation is used by such tools in practice, in order to
+improve scalability. The obvious advantage
+of this approach is that it works directly on the implementation. However, most
+static analysis tools aim at discovering 
+more general implementation deficiencies, such as uninitialized variables, null
+pointer references, or out of bound array indices. An additional drawback
+is their 
+practical availability for languages other than C and Java. To the best of our
+knowledge, no such tool exists for Python.
 Both runtime verification and static analysis can be seen as additions to
 traditional model checking, rather than alternatives. Some progress has been 
-made in the direction of combining the approaches~\cite{Java_PathFinder,Linux_driver,Brat01combiningstatic}. For instance, Java PathFinder~\cite{Java_PathFinder}
-relies on static analysis to reduce the state space before systematically exploring it by executing the abstracted program. In~\cite{Brat01combiningstatic}
-static analysis and model checking also operate in a feedback loop. The drawback of static analyzers is their 
-practical availability for languages other than C and Java. To the best of our knowledge, no such tool exists for Python.
-However, our approach of using code slicing to abstract the implementation shares similarities.
-
-TODO: trace mining
-
+made in the direction of combining the
+approaches~\cite{Java_PathFinder,Linux_driver,Brat01combiningstatic}. For
+instance, Java PathFinder~\cite{Java_PathFinder}
+relies on static analysis to reduce the state space before systematically
+exploring it by executing the abstracted program.
+In~\cite{Brat01combiningstatic}
+static analysis and model checking also operate in a feedback loop.
+
+Instead of constructing a model of the system from scratch, when the only 
+source of documentation is the code itself, a technique called process (or workflow) mining~\cite{Agrawal:1998:MPM:645338.650397}
+can be used. It is a method of automatically distilling a process or software specification from a set of 
+structured traces collected from real system executions.
+The mined specifications are typically expressed in a form of Message Sequence Graphs or Petri-Nets, 
+graphic formalism variants of the classical finite-state machines. Given that the 
+analysis of program execution traces is still the most prevalent debugging technique
+for distributed systems, and DIRAC's components (as well as the middleware) produce
+immense amounts of execution logs anyhow, it seems like an attractive approach to constructing models.
+There is a vast amount of work in this area~\cite{Aalst:1997:VWN:647744.733919,Lou:2010:MPW:1835804.1835883,Walkinshaw:2008:IFM:1642931.1642962,vanderAalst:2004:WMD:1018031.1018329,Ammons:2002:MS:503272.503275,Kumar:2011:SMC:1985793.1986002}.
+A common assumption in process mining is that each trace is totally ordered, 
+which for concurrent systems does not always hold. To cope with this, most approaches 
+require that the thread IDs or process IDs are recorded in the traces as well.
+Observing this as a realistic obstacle for obtaining an accurate model, Lou et al. in~\cite{Lou:2010:MPW:1835804.1835883}
+propose an algorithm for constructing models from traces of concurrent systems by expressing events as partially ordered.
+Their algorithm identifies dependency relationships between pairs of events in interleaved
+traces. Another common assumption is a direct result of the underlying techniques used for constructing the models, namely data mining 
+and machine learning. These techniques consider ``erroneous activities'' or activities that
+were not logged for some reason, as noise, and disregard trace sequences with no statistical significance.
+As a result, a design flaw can be mistaken for a noise, or not captured at all if a particular 
+execution path has a low probability of occurrence. A notable exception in this direction is~\cite{vanderAalst:2005:PMV:2099016.2099032}, where the emphasis is not only on process discovery,
+but also on verification: desired or undesired properties can be expressed with Linear Temporal Logic, 
+and the checker verifies whether the observed behavior in the traces matches the property.
+
+While relying solely on process mining as a technique for extracting a formal model may not be 
+the way to go, inferring a ``fuzzy'' model from the frequent interaction traces, rather than manual code analysis
+could be a helpful initial step. Nevertheless, the graphical models obtained must be 
+translated into a process algebra formalism accepted by an actual toolset, if system correctness is to be established.
 \section{Conclusions}

### 2013-02-11 ElsevierFGCS_singleColumn.tex
library cleanup

@@
 variants to explicit model checking have gained popularity. Runtime
-verification~\cite{citeulike:10198274,Havelund:2008:RVC:1424465.1424470}
+verification~\cite{DBLP:conf/dagstuhl/ColinM04,Havelund:2008:RVC:1424465.1424470}
 aims to increase confidence in the correctness, but does not make claims for
@@
 immense amounts of execution logs anyhow, it seems like an attractive approach to constructing models.
-There is a vast amount of work in this area~\cite{Aalst:1997:VWN:647744.733919,Lou:2010:MPW:1835804.1835883,Walkinshaw:2008:IFM:1642931.1642962,vanderAalst:2004:WMD:1018031.1018329,Ammons:2002:MS:503272.503275,Kumar:2011:SMC:1985793.1986002}.
-A common assumption in process mining is that each trace is totally ordered, 
+There is a vast amount of work in this area~\cite{Aalst:1997:VWN:647744.733919,Lou:2010:MPW:1835804.1835883,Walkinshaw:2008:IFM:1642931.1642962,vanderAalst:2004:WMD:1018031.1018329,Ammons:2002:MS:503272.503275}.
+A common assumption in process mining is that the events in each trace are totally ordered, 
 which for concurrent systems does not always hold. To cope with this, most approaches 
@@
 and machine learning. These techniques consider ``erroneous activities'' or activities that
-were not logged for some reason, as noise, and disregard trace sequences with no statistical significance.
+were not logged in the appropriate order for some reason, as noise, and disregard trace sequences with no statistical significance.
 As a result, a design flaw can be mistaken for a noise, or not captured at all if a particular 

### 2013-02-12 ElsevierFGCS_singleColumn.tex
deductive verification added to related work.

@@
 practical availability for languages other than C and Java. To the best of our
-knowledge, no such tool exists for Python.
-Both runtime verification and static analysis can be seen as additions to
+knowledge, no such tool exists for Python. Both runtime verification and static analysis can be seen as additions to
 traditional model checking, rather than alternatives. Some progress has been 
@@
 static analysis and model checking also operate in a feedback loop.
+
+Static analysis is, in a sense, an extreme form of automated deductive
+verification.  In its original form, deductive verification technology
+allowed for the use of logics such as Floyd-Hoare logic~\cite{Hoa:69} for
+showing the correctness of a software or hardware system by constructing
+proofs thereof, based on mathematical descriptions of the behavior of
+the system.  These days, the proofs involved are often discharged using a
+mechanical or an automated theorem prover. Moreover, more advanced logics
+such as Separation Logic~\cite{Rey:02} are used. Deductive verification
+is applied quite successfully on software code~\cite{PMSJP:12,KAE+:10}
+and hardware~\cite{AATH:10}, but the effort required for conducting the
+verification can be prohibiting, and can compare poorly to model checking.
 

### 2013-02-12 ElsevierFGCS_singleColumn.tex
Introduction: added a few lines on the methodology for performing the check, and contributions. Also on the loss-of-efficiency statistics. Conclusions: added a few lines on the general applicability of the approach on other systems.

@@
 and the distributed knowledge of different subsystems within the collaboration.
+Depending on the frequency of certain heavier data (re)processing campaigns throughout 
+the year, the occurrence of such inconsistencies can be up to several reported cases 
+per month. However, since they have a blocking effect on these heavy grid activities
+with a strict deadline, locating and fixing the root cause of such a problem 
+typically has a lower priority within the collaboration, compared to a temporary fix, 
+such as manipulating database records or restarting processes manually.
 
@@
 others unveiled more elementary design problems. We further formulated,
-formalized and verified several application-specific properties.
+formalized and verified several application-specific properties. Once
+the model is in place, the toolset can automatically check for general system
+properties, such as deadlocks. We show how to formulate and check
+application-specific
+properties with the toolset. Furthermore, we outline a methodology to
+effectively tackle the classical state space explosion
+problem arising when the underlying model grows beyond the resources available
+to perform explicit model checking. For this, we borrow 
+concepts from finite state machine formalisms for description of system properties.
+Extending the specification with a so-called monitoring automaton for tracking 
+relevant transitions in the original model, can speed up the
+analysis and discovery of traces that lead to violation of a certain property.
+This idea is simple but flexible and generic enough to describe many temporal properties of concurrent programs.
 
@@
 and modeling the behavior of these subsystems. Section 4 presents the
-analysis with the mCRL2 toolset and the issues detected. Section 5 and 6
+analysis with the mCRL2 toolset and the issues detected. Work related to the 
+domain of distributed systems verification is discussed in Section 5. Section 6 and 7
 conclude and give directions for future work, respectively.
@@
 analysis. 
-It provides a concurrent use of over 60K CPUs and 10M file replicas distributed over
-tens of grid sites, provided by the WLCG computing grid \cite{Bird:2011zz}. The community of users has grown to several hundreds, loading the 
-LHC grid resources with up to 30K simultaneously running jobs (Fig.~\ref{fig:LHCb-resource-usage}, top) and over 100K jobs executed per day, during peak processing periods.
-About 10PB of data is stored and visible in the LHCb file catalogs, with daily data transfers between storage units reaching 3Gbps (Fig.~\ref{fig:LHCb-resource-usage}, bottom).
-\begin{figure*}[tp!]
+It provides a concurrent use of over 60K CPUs and 10M file replicas distributed
+over
+tens of grid sites, provided by the WLCG computing grid \cite{Bird:2011zz}. The
+community of users has grown to several hundreds, loading the 
+LHC grid resources with up to 30K simultaneously running jobs
+(Fig.~\ref{fig:LHCb-resource-usage}, top) and over 100K jobs executed per day,
+during peak processing periods.
+About 10PB of data is stored and visible in the LHCb file catalogs, with daily
+data transfers between storage units reaching 3Gbps
+(Fig.~\ref{fig:LHCb-resource-usage}, bottom).
+\begin{figure*}
   \centering
@@
 \section{System modeling}
-
+\label{section3}
 \subsection{The mCRL2 language and toolset}
@@
 static analysis tools aim at discovering 
-more general implementation deficiencies, such as uninitialized variables, null
-pointer references, or out of bound array indices. An additional drawback
+more general and low-level implementation bugs, such as uninitialized variables, null
+pointer references, zero division, or out of bound array indices. 
+It is almost impossible for a general static analysis tool to
+figure out the exact high-level intent of the implementation code taken as input.
+Because of the heavy
+approximation they use,
+static analysis tools are quite fast and applicable to large software projects,
+but result in many false positives reported. 
+An additional drawback
 is their 
 practical availability for languages other than C and Java. To the best of our
-knowledge, no such tool exists for Python. Both runtime verification and static analysis can be seen as additions to
+knowledge, no such tool exists for Python. Both runtime verification and static
+analysis can be seen as additions to
 traditional model checking, rather than alternatives. Some progress has been 
@@
 components of the same kind deliver the overall system functionality. 
-Once becoming proficient with
+It is observed that many large concurrent/distributed systems are essentially
+parametrized, meaning that they contain components or process descriptions
+which are behaviorally similar. Once becoming proficient with
 the formal language notation, modeling such components is
-reduced to reapplying similar abstraction rules.
+reduced to reapplying similar abstraction rules. The methodology we have outlined is particularly
+suitable for such systems. In addition, the guidelines proposed can be applied
+when dealing with data-centered concurrent systems, where the only documentation
+is the implementation itself, and most of the problems arise from sharing data among 
+processes. The biggest challenge for such systems is finding a reasonable abstraction
+level for the formal model, and we believe that the combination of code slicing and
+abstraction rules of thumb presented in Section 3 can be helpful in this direction.
+
 Practitioners

### 2013-02-14 FGCS_rebuttal.txt
We're almost ready to go.

We value the effort invested by the reviewers in assessing our paper.
A list of changes for the raised points:
-------------
* [Introduction clearly describes one of the main interests of the paper: a methodology for building models (in the model-checking sense) for a legacy system. On the other hands, it fails to introduce the second contribution of the paper, which is a methodology to actually performing the check...]

> Introduction has been modified to introduce these aspects, paragraph:
[Once the model is in place, the toolset can automatically check for general system properties, such as deadlocks. We show how to formulate and check application-specific properties... 
This idea is simple but flexible and generic enough to describe many temporal properties of concurrent systems.]
-------------
* [However, for a reader not familiar with the history of the EGEE/EGI/WLCG grid, some information is missing. Mainly, Dirac appears as a standalone system, and the underlying software and hardware infrastructure is not described.
- The paper should explain the relation of Dirac with the gLite middleware, and provide a reference to  glite (eg E. Laure, S.M. Fisher, A. Frohner, C. Grandi, P. Kunszt, et al. Programming the Grid with gLite. Computational Methods in Science and Technology, 12(1):33-45, 2006.).]

> This explanation has been added in Section 2.2, paragraph:
[Originally, the DIRAC WMS submitted pilot jobs using the native gLite middleware tools, but this method was shown to be suboptimal. The gLite resource brokers... making sites compete with each other for jobs and have a faster job turnaround.]
> gLite reference also added
-------------
* [The paper should give figures showing the size and scale of distribution of the hardware infrastructure (number of sites, of CPUs or cores, storage capacity etc.) accessible to DIRAC.]
> Paragraph added:
[It provides a concurrent use of over 60K CPUs and 10M file replicas distributed over tens of grid sites, ....About 10PB of data is stored and visible in the LHCb file catalogs, with daily data transfers between storage units reaching 3Gbps.] 
> Figures 1 and 2 added for illustration.
-------------
* [Considering the SMS, a state  machine analogous to the one for WMS would improve the readability of section 2.3.]
> Figure 7 added.
-------------
* [ Other approaches of model checking (eg deductive verification) could have been discussed. Also, the choice of model checking versus trace mining (a very active research area) should be justified, as both Dirac and gLite services generate copious logs that were not exploited for diagnosis. Overall, a Related Work section is necessary.]
> Related work was addressed by all reviewers, so we dedicated a separate Section 5 for this.

* [Section 4 is also very pedagogical: it follows the real-world checking process, including preliminary visualization. However, figures 7 should be provided with a better resolution and better explained ... Figure 8 should probably printed in color, or entirely removed. If kept, the "suspicious transition" should be pointed in the figure.]
> Figure 7 (now 10) is now with a better resolution (background circles simply indicate which portion of the entire state space belongs to this self-transitions. We found it un-necessary to bother the reader with this detail). Figure 8 (now 11): suspicious transition is clearly pointed at, now.
-------------
* [Section 4.3 demonstrates the efficiency of the approach, on the smallest of the two state machine. On the other hand, the approach fails on the WMS, due to combinatorial explosion.  The strategy to cope with this issue is presented as a "trick", and should be better related to the relevant concepts in the literature, in particular to testing.]
> We briefly relate this in Section 5. Related Work [This component resembles our monitoring process automaton used to cope with state space explosion during model checking of DIRAC’s WMS system]
-------------
* [The authors motivate the problem of lost efficiency of resource usage and manpower. Are there any statistics showing actual loss of efficiency? How often the DIRAC system get into inconsistent state?]
> Added in Introduction: [Depending on the frequency of certain heavier data (re)processing campaigns throughout the year, the occurrence of such inconsistencies can be up to several reported cases per month..., compared to a temporary fix, like manipulating database records or restarting processes manually]
-------------
* [The application of model checking on DIRAC system and finding race conditions is useful. However, the extendability of this approach for other systems is not clear to me. Can these approaches be applied on other systems? What changes need to be done?]
> We attempt to tackle this in Section 6. Conclusions. A few more lines of explanation are added in: [It is observed that many large concurrent/distributed systems are essentially parametrized, meaning that they contain components or process descriptions which are behaviorally similar...and we believe that the combination of code slicing and abstraction rules of thumb presented in Section 3 can be helpful in this direction.]
-------------
* Reviewer #4's comments again raise the issues of applicability of the approach to other systems, and in depth discussion of related work. (covered above)
-------------
* [Some discussion on the complexity and/or time requirements of using the model would be interesting to understand the usefulness of the model]
> This is difficult to generalize for any model, but we provide statistics on the state space generation and model checking times in section 4.Analysis and Issues.

Best regards,
Daniela

### 2013-02-14 ElsevierFGCS_singleColumn.tex
We're almost ready to go.

@@
 typically has a lower priority within the collaboration, compared to a temporary fix, 
-such as manipulating database records or restarting processes manually.
+like manipulating database records or restarting processes manually.
 
@@
 to perform explicit model checking. For this, we borrow 
-concepts from finite state machine formalisms for description of system properties.
+concepts from finite state machine formalisms for description of the system properties.
 Extending the specification with a so-called monitoring automaton for tracking 
@@
 analysis and discovery of traces that lead to violation of a certain property.
-This idea is simple but flexible and generic enough to describe many temporal properties of concurrent programs.
+This idea is simple but flexible and generic enough to describe many temporal properties of concurrent systems.
 
@@
 
+\begin{figure}[tp!]
+\includegraphics[width=0.7\textwidth]{./graphics/Figure1_1.png}
+\centering
+\caption{LHCb job troughput by site}
+\label{fig:Job-throughput}
+\end{figure}
+\begin{figure}[tp!]
+\includegraphics[width=0.7\textwidth]{./graphics/Figure1_2.png}
+\centering
+\caption{LHCb data transfers throughput}
+\label{fig:Data-transfers}
+\end{figure}
 The development of DIRAC started in 2002 as a system for production of simulation
@@
 LHC grid resources with up to 30K simultaneously running jobs
-(Fig.~\ref{fig:LHCb-resource-usage}, top) and over 100K jobs executed per day,
+(Fig.~\ref{fig:Job-throughput}) and over 100K jobs executed per day,
 during peak processing periods.
@@
 data transfers between storage units reaching 3Gbps
-(Fig.~\ref{fig:LHCb-resource-usage}, bottom).
-\begin{figure*}
-  \centering
-  {\label{fig:Job-throughput}\includegraphics[width=0.7\linewidth]{./graphics/Figure1_1.png}}    
-  \vfill
-  \vspace{14pt}
-{\label{fig:Data-transfers}\includegraphics[width =0.7\linewidth]{./graphics/Figure1_2.png}}
-  \caption{LHCb resource usage}
-  \label{fig:LHCb-resource-usage}
-\end{figure*}
+(Fig.~\ref{fig:Data-transfers}).
+
 Python was chosen as the implementation language, since it enables
@@
 Fig.~\ref{fig:DIRAC-Arch}.
-
-\begin{figure}[tp]
-\includegraphics[width=0.5\textwidth] {./graphics/Figure1.png}
-\centering
-\caption{DIRAC Architecture overview}
-\label{fig:DIRAC-Arch}
-\end{figure}
 
@@
 entities.
-
+\begin{figure}[tp!]
+\includegraphics[width=0.5\textwidth] {./graphics/Figure1.png}
+\centering
+\caption{DIRAC Architecture overview}
+\label{fig:DIRAC-Arch}
+\end{figure}
 \textit{Resources} are software abstractions of the underlying heterogeneous grid
@@
 
-The processing begins with the \textit{Request Preparation Agent}. It selects all the
+Fig.~\ref{fig:CacheReplicas-SM} depicts the state machine of the CacheReplicas entitiy. The processing begins with the \textit{Request Preparation Agent}. It selects all the
 ``New'' replica entries, checks whether they are registered in a Logical
@@
 of the corresponding CacheReplicas to "Failed'' and ``Waiting'' respectively.
+
 \definecolor{light-gray}{gray}{0.90}
@@
 \subsection{Simulation and debugging}
+\begin{figure*}[tp!]
+  \centering
+  \subfloat[Logging info of a DIRAC job]{\label{fig:Job-Logging-Info}\includegraphics[width=0.4566\linewidth]{./graphics/Figure6a.png}}    
+  \hfill
+  \subfloat[XSim simulator trace]{\label{fig:simulator}\includegraphics[width =0.5\linewidth]{./graphics/Figure6b.png}}
+  \caption{Invalid job state transitions}
+  \label{fig:Simulator-job-trace}
+\vspace{-8 pt}
+\end{figure*}
 Apart from verification, the mCRL2 toolset has a rich set of tools for analysis of the modeled system. 
@@
 before they actually surface in the real system.
-\begin{figure*}[bp!]
-  \centering
-  \subfloat[Logging info of a DIRAC job]{\label{fig:Job-Logging-Info}\includegraphics[width=0.4566\linewidth]{./graphics/Figure6a.png}}    
-  \hfill
-  \subfloat[XSim simulator trace]{\label{fig:simulator}\includegraphics[width =0.5\linewidth]{./graphics/Figure6b.png}}
-  \caption{Invalid job state transitions}
-  \label{fig:Simulator-job-trace}
-\vspace{-8 pt}
-\end{figure*}
 One such problematic behavior was recently reported in DIRAC's production (Fig.~\ref{fig:Job-Logging-Info}), where
@@
 \subsection{Visualization}
-\begin{figure}[tp]
-\vspace{-10 pt}
-\includegraphics[width=0.5\linewidth,keepaspectratio=true]{./graphics/Figure7_new.png}
-\centering
-\caption{CacheReplicaMem process visualization with DiaGraphica}
-\label{fig:DiaGraphica}
-\end{figure}%123478
 Reasonably small LTSs can be easily visualized with the interactive GUI tools,
@@
 on the CacheReplicas memory process, which resulted in only 6 interesting states for this process alone.
-
+\begin{figure}[tp]
+\vspace{-10 pt}
+\includegraphics[width=0.5\linewidth,keepaspectratio=true]{./graphics/Figure7_new.png}
+\centering
+\caption{CacheReplicaMem process visualization with DiaGraphica}
+\label{fig:DiaGraphica}
+\end{figure}%123478
 The clustering
@@
 instead of hardcoding states in SQL queries can prevent such bugs.
-\begin{figure}[bp!]
+\begin{figure}[tp!]
 \vspace{-10 pt}
@@
 function calls and variable assignments, 
-in order to capture and emit relevant events which are then monitored against a
+in order to capture and emit relevant events which are then monitored by a
 special component during execution.
@@
 of DIRAC's WMS system. While runtime verification is useful for
-partially-capturing incorrect behavior of the system implementation, 
+capturing incorrect behavior of the system implementation, 
 code instrumentation is not always a feasible option, considering the increase
@@
 such as Separation Logic~\cite{Rey:02} are used. Deductive verification
-is applied quite successfully on software code~\cite{PMSJP:12,KAE+:10}
+is applied quite successfully on software~\cite{PMSJP:12,KAE+:10}
 and hardware~\cite{AATH:10}, but the effort required for conducting the
@@
 describe the system behavior in the abstract model. With this critical step in mind, formal methods
-are an important addition to the assembly of software analysis and verification techniques for concurrent distributed systems.
+are an important addition to the assembly of software analysis and verification techniques for concurrent systems.
 \section{Future work}

### 2013-02-15 FGCS_rebuttal.txt
..

@@
 
-> Introduction has been modified to introduce these aspects, paragraph:
+> Introduction has been extended, paragraph:
 [Once the model is in place, the toolset can automatically check for general system properties, such as deadlocks. We show how to formulate and check application-specific properties... 
@@
 If kept, the "suspicious transition" should be pointed in the figure.]
-> Figure 7 (now 10) is now with a better resolution (background circles simply indicate which portion of the entire state space belongs to this self-transitions.
+> Figure 7 (now Figure 10) is now with a better resolution (background circles are removed.
+They simply indicate which portion of the entire state space belongs to this self-transitions.
 We found it un-necessary to bother the reader with this detail).
-Figure 8 (now 11): suspicious transition is clearly pointed at, now.
+Figure 8 (now Figure 11): suspicious transition is clearly marked now.
 -------------
@@
 Conclusions.
-A few more lines of explanation are added in: [It is observed that many large concurrent/distributed systems are essentially parametrized, meaning that they contain components or process descriptions which are behaviorally similar...and we believe that the combination of code slicing and abstraction rules of thumb presented in Section 3 can be helpful in this direction.]
+A few more lines of explanation are added: [It is observed that many large concurrent/distributed systems are essentially parametrized, meaning that they contain components or process descriptions which are behaviorally similar...and we believe that the combination of code slicing and abstraction rules of thumb presented in Section 3 can be helpful in this direction.]
 -------------
-* Reviewer #4's comments again raise the issues of applicability of the approach to other systems, and in depth discussion of related work. (covered above)
+* Reviewer #4's comments again raise the issues of applicability of the approach to other systems, and more in depth discussion of related work. (covered above)
 -------------
@@
 
-Best regards,
-Daniela
+Best regards

### 2013-02-15 ElsevierFGCS_singleColumn.tex
..

@@
 
-\begin{figure}[tp!]
-\includegraphics[width=0.7\textwidth]{./graphics/Figure1_1.png}
+\begin{figure}[bp!]
+\includegraphics[width=0.7\textwidth]{./graphics/Figure1.png}
 \centering
@@
 \begin{figure}[tp!]
-\includegraphics[width=0.7\textwidth]{./graphics/Figure1_2.png}
+\includegraphics[width=0.7\textwidth]{./graphics/Figure2.png}
 \centering
@@
 entities.
-\begin{figure}[tp!]
-\includegraphics[width=0.5\textwidth] {./graphics/Figure1.png}
+\begin{figure}[bp!]
+\includegraphics[width=0.5\textwidth] {./graphics/Figure3.png}
 \centering
@@
 \begin{figure}[bp]
-\includegraphics[width=0.7\textwidth]{./graphics/Figure2.png}
+\includegraphics[width=0.7\textwidth]{./graphics/Figure4.png}
 \centering
@@
 \begin{figure}[tp]
-\includegraphics[width=0.7\linewidth,keepaspectratio=true]{./graphics/Figure3.png}
+\includegraphics[width=0.7\linewidth,keepaspectratio=true]{./graphics/Figure5.png}
 \centering
@@
   \centering
-  \includegraphics[width=0.98\linewidth]{./graphics/Figure4.png}
+  \includegraphics[width=0.98\linewidth]{./graphics/Figure6.png}
   \caption{DIRAC Storage Management System}
@@
   \centering
-  \includegraphics[width=0.95\linewidth]{./graphics/SMS_CacheReplicasSM.png}
+  \includegraphics[width=0.95\linewidth]{./graphics/Figure7.png}
   \caption{CacheReplicas state machine}
@@
 Abstraction aims at reducing
-the program's state space in order to overcome the resource limitations \cite{Pelánek08fightingstate} encountered during model-checking. 
+the program's state space in order to overcome the resource limitations \cite{Pelanek08fightingstate} encountered during model-checking. 
 Furthermore, specification languages describe \textit{what} is being done, 
@@
 \begin{center}
- \includegraphics[width=0.5\linewidth,keepaspectratio=true]{./graphics/Figure5.png}
+ \includegraphics[width=0.5\linewidth,keepaspectratio=true]{./graphics/Figure8.png}
 \end{center}
@@
   \centering
-  \subfloat[Logging info of a DIRAC job]{\label{fig:Job-Logging-Info}\includegraphics[width=0.4566\linewidth]{./graphics/Figure6a.png}}    
+  \subfloat[Logging info of a DIRAC job]{\label{fig:Job-Logging-Info}\includegraphics[width=0.4566\linewidth]{./graphics/Figure9a.png}}    
   \hfill
-  \subfloat[XSim simulator trace]{\label{fig:simulator}\includegraphics[width =0.5\linewidth]{./graphics/Figure6b.png}}
+  \subfloat[XSim simulator trace]{\label{fig:simulator}\includegraphics[width =0.5\linewidth]{./graphics/Figure9b.png}}
   \caption{Invalid job state transitions}
@@
 \vspace{-10 pt}
-\includegraphics[width=0.5\linewidth,keepaspectratio=true]{./graphics/Figure7_new.png}
+\includegraphics[width=0.5\linewidth,keepaspectratio=true]{./graphics/Figure10.png}
 \centering
@@
 \vspace{-10 pt}
-\includegraphics[width=0.57\linewidth,keepaspectratio=true]{./graphics/Figure8.png}
+\includegraphics[width=0.57\linewidth,keepaspectratio=true]{./graphics/Figure11.png}
 \centering
@@
 \begin{figure}[tp]
-\includegraphics[width=0.7\linewidth,keepaspectratio=true]{./graphics/Figure9.png}
+\includegraphics[width=0.7\linewidth,keepaspectratio=true]{./graphics/Figure12.png}
 %was PropertyViolation1.png
@@
 \begin{figure}[tp]
-\includegraphics[width=0.5\linewidth,keepaspectratio=true]{./graphics/Figure10.png}
+\includegraphics[width=0.5\linewidth,keepaspectratio=true]{./graphics/Figure13.png}
 \centering
@@
 \textit{error} is executed, and can be detected by the LTSmin tool. 
-\begin{figure}[bp]
-\includegraphics[width=0.6\linewidth,keepaspectratio=true]{./graphics/Figure11.png}
+\begin{figure}[tp]
+\includegraphics[width=0.6\linewidth,keepaspectratio=true]{./graphics/Figure14.png}
 \centering
@@
 \begin{figure}[tp]
-\includegraphics[width=0.6\linewidth,keepaspectratio=true]{./graphics/Figure12.png}
+\includegraphics[width=0.6\linewidth,keepaspectratio=true]{./graphics/Figure15.png}
 \centering
